Circadian regulation of endoplasmic reticulum calcium response in mouse cultured astrocytes

Version 1

Abstract

Abstract

The circadian clock, an internal time-keeping system orchestrates 24-hour rhythms in physiology and behavior by governing rhythmic transcription within cells. Astrocyte, the most abundant glial cell type, play crucial roles in central nervous system functions. However, a detailed understanding of how the circadian clock impacts functions of astrocyte remains largely unexplored. In this study, utilizing circadian clock-synchronized mouse cultured cortical astrocytes and RNA sequencing, we identified 412 circadian rhythmic transcripts with a distinct astrocyte-specific expression pattern. A Gene Ontology analysis of these rhythmic transcripts highlighted genes implicated in Ca2+ homeostasis as being under circadian control. Notably, Herpud1 (Herp) exhibited robust circadian rhythmicity at both mRNA and protein levels, a rhythm disrupted in astrocytes lacking the circadian transcription factor, BMAL1. HERP regulated endoplasmic reticulum (ER) Ca2+ release by modulating the degradation of inositol 1,4,5-trisphosphate receptors (ITPRs). Intriguingly, ATP-stimulated ER Ca2+ release varied with the circadian cycle, being more pronounced at subjective night, likely owing to the rhythmic expression of ITPR2. Furthermore, this rhythmic ER Ca2+ response led to day/night variations in the phosphorylation of Cx43 (Ser368) and the gap junctional communication. Given the role of gap junction channel (GJC) in propagating Ca2+ signals, we suggest that this circadian regulation of ER Ca2+ responses could markedly affect astrocytic modulation of synaptic activity according to the time of day. Overall, our study enhances the understanding of how circadian clock influences astrocyte function in the CNS, shedding light on their potential role in daily variations of brain activity and health.

Introduction

The circadian clock orchestrates a range of behaviors, physiological processes, and cellular and biochemical activities in organisms with approximately 24-hour rhythms, facilitating adaptation to environmental changes (Bell-Pedersen et al., 2005). In mammals, the master clock localized in the suprachiasmatic nucleus (SCN) of the anterior hypothalamus, synchronizes peripheral tissue clocks throughout the body to the light/dark cycle (Mohawk et al., 2012). The circadian clock’s molecular mechanism is driven by a transcriptional translational feedback loop (TTFL). CLOCK and BMAL1 form a heterodimer that triggers expression of the repressor genes, Period (Per) and Cryptochrome (Cry), in addition to other clock-controlled output genes. PER and CRY proteins then translocate to the nucleus, where they interact with CLOCK/BMAL1, halting further transcriptional activation. A secondary feedback loop involving the interplay between transcriptional activation by ROR and repression by REV-ERB finely regulates the rhythmic transcription of Bmal1 (Takahashi, 2017). Recent studies have revealed that the circadian clock controls the rhythmic transcription of a large number of genes, displaying notable tissue specificity. This characteristic underscores the time-of-day-specific functional specialization of diverse tissues (Mure et al., 2018; Zhang et al., 2014). Consequently, circadian clock-controlled genes significantly influence various physiological processes including development, metabolism, aging, and neurodegeneration (Lananna & Musiek, 2020; Reinke & Asher, 2019; Schultz & Kay, 2003).

Astrocytes, a type of glial cell and the most abundant cell type in the CNS play vital roles in CNS function (Sofroniew & Vinters, 2010). They maintain the homeostasis of extracellular fluids, ions, and neurotransmitters; provide energy substrates to neurons ; modulate local blood flow ; assist in interstitial fluid drainage. Beyond these roles, astrocytes interact functionally with brain cells including neurons and astrocytes, contributing actively to brain circuitry (Verkhratsky & Nedergaard, 2018). Though not electrically excitable, astrocytes exhibit complex intracellular Ca2+ signaling, both spatially and temporally, within individual astrocytes and across astrocytic networks, thereby influencing synaptic activity (Semyanov et al., 2020). Activation of Gq-protein coupled receptor (GPCR) by external stimuli initiates the inositol 1,4,5-trisphosphate receptor (IP3R) signaling cascade, causing a release of Ca2+ from the ER that substantially increases intracellular Ca2+ levels (Guerra-Gomes et al., 2017; Verkhratsky & Nedergaard, 2018). Astrocytes express various types of Gq-coupled receptors activated by neurotransmitters such as ATP, epinephrine, glutamate, and dopamine etc. (Verkhratsky & Nedergaard, 2018). Increases in intracellular Ca2+ in astrocytes lead to alterations in cell-to-cell communication and/or induce gliotransmission (Giaume et al., 2021; Verkhratsky & Nedergaard, 2018). Recent studies have shown that astrocyte Ca2+ signaling can regulate organism-level behaviors, such as circadian rhythms and sleep patterns (Bojarskaite et al., 2020; Brancaccio et al., 2019) as well.

Accumulating evidence underscores the importance of the circadian clock in astrocytes function. First of all, cultured murine cortical astrocytes expressing a PER-luciferase reporter (PER::LUC) exhibit circadian oscillations per expression (Prolo et al., 2005). These astrocytes can be entrained by various cues, including temperature cycle (Prolo et al., 2005) and vasoactive intestinal polypeptide (VIP), which is the neuropeptide synchronizing SCN (Marpegan et al., 2009). These findings provide strong evidence that a molecular clock operates in astrocytes. In addition, disruption of clock genes specific to astrocytes has been linked to a range of deleterious effects, including excessive astrocyte activation, disruptions in daily activity patterns, impaired learning, metabolic imbalances, and a shortened lifespan (Barca-Mayo et al., 2020; Barca-Mayo et al., 2017; Griffin et al., 2019; Lananna et al., 2018). Remarkably, these phenotypic changes closely resemble those observed in astrocytes during the progression of neurodegenerative diseases (Brandebura et al., 2023; Phatnani & Maniatis, 2015). Moreover, the rhythmic metabolic activity of astrocytes, including intracellular Ca2+ fluctuations and neurotransmitter release, is critical for circadian timing in the SCN (Brancaccio et al., 2017). Despite the emerging significance of circadian rhythms in astrocyte function in health and disease, our understanding of the link between the circadian clock and specific astrocytic functions is still limited.

In this study, we conducted a transcriptome analysis to investigate how the circadian clock regulates astrocyte function. Our research uncovered that Herp (homocysteine-inducible ER protein with ubiquitin like domain 1; Herp) exhibited a BMAL1-dependent rhythmic expression pattern, peaking at midday and reaching its nadir at mid-night. Herp1 was important for ER Ca2+ regulation in cultured astrocytes. ATP stimulated ER Ca2+ response varied according to circadian time (CT), a dependence that was absent in cultured astrocytes from Bmal1-/- mice. This response was mediated by HERP1-mediated degradation of the inositol 1,4,5-triphosphate receptor 2 (ITPR2) degradation. Furthermore, the pronounced release of Ca2+ during the subjective night led to robust phosphorylation of the Ser368 residue of CONNEXIN 43 (CX43), likely diminishing cell-to-cell communication between astrocytes. Overall, our study demonstrates that the circadian clock orchestrates a variety of astrocytic processes including Ca2+ homeostasis by controlling the rhythmic transcription of genes, highlighting its pivotal role in CNS function.

Results

Defining circadian rhythmic transcripts in mouse cultured cortical astrocytes

To understand how the circadian clock regulates astrocyte physiology, we conducted a circadian transcriptome analysis in cultured mouse cortical astrocytes. Cultured astrocytes were subjected to serum shock (SS), which synchronizes the circadian clock, and then harvested at 4 hour intervals for 2 days beginning 12hrs after. A time series astrocyte transcriptome was determined RNA-seq (Fig. 1A). The Gaussian mixture model was used to define the sets of expressed transcripts. We chose 0.577 transcripts per million (TPM) as a cut-off, corresponding to the 1% threshold in the distribution curve of highly expressed genes (Fig. S1A). A total of 17,671 transcripts, with the highest value among the 12 time points exceeding the TPM threshold 0.577, were considered to be expressed. To validate astrocyte enrichment in our cell culture system, we compared the expression of marker genes for nervous cells (Fig. S1B). Time-merged mean TPM values indicated that transcriptome data were replete with astrocyte markers but lacked markers for microglia, oligodendrocytes, neurons, and endothelial cell.

Circadian rhythmic transcripts in mouse cultured cortical astrocytes.

(A) Experimental scheme for synchronizing circadian rhythms in mouse cultured cortical astrocytes culture with subsequent RNA-sequencing performed over 2 days. (B) Venn diagram displaying the number of circadian rhythmic transcripts identified by two algorithms (q < 0.05 in MetaCycle or BioCycle). (E) Expression profiles of Plat, Il34, Adora1, and Mybph in SS-synchronized cultured astrocytes obtained from RNA-seq data. The q-values for BioCycle (red) and MetaCycle (blue) are displayed in the bottom right corner. (C) Radial histogram of the distribution of phases of rhythmic genes in the astrocyte transcriptome. TASS was transformed to CT for ease of interpretation (TASS 8 hour = CT0). (D) Top 10 enriched GO Biological Process (BP) terms for significant astrocyte circadian rhythmic genes (p < 0.01) identified by Metascape (https://metascape.org). See also S1-S3 and Table S1.

To assess rhythmically oscillating genes in time-series transcriptome datasets, we applied BioCycle and MetaCycle packages, which have had been successfully used for this purpose (Agostinelli et al., 2016; Wu et al., 2016), to the astrocyte transcriptome. BioCycle and MetaCycle analyses detected 321 and 311 periodic transcripts, respectively (FDR corrected, q-value < 0.05) (Fig. 1B). Among those transcripts, 220 (53.4%) were detected by both methods, but many transcripts were not overlapping. For example, Plat and Il34 were detected as rhythmic only in BioCycle and Mybph and Adora1 were detected rhythmic only in MetaCycle. Despite these incongruences, these four genes clearly exhibited a circadian rhythmic expression profile (Fig. S1C). Accordingly, we applied the union of two lists, identifying a total of 412 transcripts representing 2.3% of all transcripts as circadian rhythmic genes (Dataset S1). Oscillations of these 412 transcripts are illustrated in the heatmap in Fig. S1D.

Next, we compared rhythmic transcripts of astrocytes with those of 12 mouse tissues derived from CircaDB (Zhang et al., 2014), a database of genes that exhibit circadian expression patterns (Fig. S2A). Majority of rhythmic transcripts in astrocyte (265 of 412, 65.3%) was not rhythmic or rhythmic in one tissue. 14 transcripts were rhythmic in more than 10 tissues; most of these genes were core clock genes except Tspan4, Tsc22d3, Wee1. These results are consistent with previous reports that circadian rhythmic expression is controlled in a tissue specific manner (Mure et al., 2018; Panda et al., 2002). A comparison of the phases of these 14 common oscillating transcripts between CircaDB and our analysis showed that all 14 transcripts exhibited a consistent time difference between CircaDB and our analysis, where 8 hours after SS corresponded to CT0 (Fig. S2B). An additional comparison of phases of 82 transcripts that overlapped in only one tissue indicated a very robust phase correlation between astrocytes and the corresponding tissue (r = 0.69, p < 0.001) (Fig. S2C). Collectively, our analysis indicates that the identified oscillating transcripts were not false positive or noises but instead were clearly associated with the circadian clock. We then analyzed the phase distribution of oscillating transcripts by performing a hierarchical clustering analysis. For more intuitive and physiological interpretation, we transformed TASS phase in our data into CT phase by setting TASS 8 to CT0. Oscillating transcripts were divided into two clusters, with cluster 1 containing 185 transcripts and cluster 2 containing 227 transcripts (Fig. S3). A circular map of peak phases across all rhythmic genes revealed two major peaks; one early night peak (CT12 ∼ CT16) and one late night or dawn peak (CT21 ∼ CT1) (Fig. 1C). Circadian transcriptome studies conducted in multiple animal tissues have indicated that the phases of each tissue are not randomly distributed but are predominantly clustered in one or two narrow temporal windows (Panda et al., 2002). Consistent with these previous reports, mouse cultured astrocytes exhibited a high degree of temporal organization. Intriguingly, there was a relatively quiescent zone during subjective daytime, a finding different from that observed for diurnal animals, which showed a quiescent zone during subjective nighttime (Mure et al., 2018). To understand the biological processes and pathways controlled by the circadian clock, we performed a Gene Ontology (GO) analysis on each cluster using Metascape (https://metascape.org) (Fig. 1D). Given that the goal of our research was to investigate processes controlled by the circadian clock, we excluded 17 core clock transcripts-Arntl, Clock, Nfil3, Npas2, Bhlhe40, Bhlhe41, Cry1, Cry2, Dbp, Elf, Nr1d1, Nr1d2, Per1, Per2, Per3, Rorc, and Tef from our GO enrichment analysis. The analysis of the remaining transcripts showed that cluster 1, containing Rhoc, Bcar1, Gpld1, Hspa5, and mmp14 etc. was enriched for the biological process “positive regulation of cell migration (GO:0030335). Time varying cell motility has been reported to be important for the circadian rhythms of wound-healing processes (Hoyle et al., 2017). Although astrocytes in the adult brain are largely non-motile under normal physiological conditions, their migratory activity is enhanced during the process of astrogliosis in the context of injury (Pekny & Pekna, 2016). Given this robust circadian rhythmic expression of genes that regulate motility, the expectation is that wound repair activity of astrocytes might vary according to the time of insult and would presumably be higher toward the end of the active period. Consistent with this, the biological processes, “regulation of cell-cell junction organization” and “cell-substrate adhesion” which are important for wound repair, were also enriched in the cluster 1. One of the biological processes enriched in cluster 2, was “calcium ion homeostasis (GO:0055074)” a term that was intriguing because of the importance of intracellular Ca2+ signaling in numerous astrocyte functions (Agulhon et al., 2008; Guerra-Gomes et al., 2017). Identified Ca2+ homeostasis-related genes included Herp, Slc4a11, Sord and Kcnh1 etc. with Herp being the most significant gene with large oscillation amplitude (Table. S1).

Circadian rhythmic expression of Herp in mouse astrocyte culture

Herp was first identified as a novel gene that exhibited altered expression in response to homocysteine treatment in human umbilical vein endothelial cells (HUVECs) as part of a study on hyperhomocysteinemia (Kokame et al., 1996). Homocysteine induces ER stress and growth arrest in HUVECs (Outinen et al., 1999). Subsequent studies have shown that Herp is strongly induced not only by homocysteine but also by drugs that cause ER stress, such as tunicamycin or thapsigargin. HERP, characterized by an N-terminal ubiquitin-like domain and present in the ER membrane, facing cytoplasmic side, has emerged as a novel target protein in the unfolded protein response (UPR) (Kokame et al., 2000). In particular, Herp is a member of the ER-associated degradation (ERAD) complex, which participates in ubiquitination and relocation of ERAD substrates, as demonstrated in several cell types (e.g. HEK 293T, MEF, NIH3T3 cells) (Leitman et al., 2014). Notably, a recent study reported that HERP is capable of modulating the ER Ca2+ response through IP3R degradation (Paredes et al., 2016; Torrealba et al., 2017).

We first examined the expression patterns of core clock genes and Herp in our cultured astrocytes. Bmal1, the positive element of the TTFL, was rhythmically expressed with a peak at CT24. Per2 and Rev-Erbα, clock genes downstream of CLOCK/BMAL1, were rhythmically expressed with peaks at CT12 and CT8, respectively (Fig. 2A). Herp showed a robust rhythmic expression pattern, with a peak at CT12, that was comparable to that of Per2 (Fig. 2A). Then, in order to verify the circadian expression of Herp, we performed quantitative real-time RT-PCR (Fig. 2B). mRNA levels of Bmal1, Rev-Erbα and Herp exhibited circadian rhythm with similar phase of RNAseq results. Next, because many genes were rhythmically transcribed but their protein levels do not exhibit rhythmic changes (Reddy et al., 2006), we examined whether the levels of HERP were rhythmic. We first validated the specificity of HERP antiserum by performing Western blot analysis of astrocytes treated with siRNA targeting Herp (Fig. 2C - 2E). Herp siRNA greatly reduced both the levels of Herp mRNA (Fig. 2C) and the intensities of specific bands in Western blot analyses (Fig. 2D and 2E). Circadian clock synchronized cultured astrocytes were sampled every 6 hours for 2days and processed for Western blot analysis of BMAL1 and HERP. The phosphorylation status of BMAL1 exhibits circadian variation, such that hyper-phosphorylated BMAL1 peaks at midday and hypo-phosphorylated BMAL1 peaks early or at the end of the day (Yoshitane et al., 2009). In our SS-synchronized cultured astrocytes, BMAL1 was hyperphosphorylated at mid-day (e.g. CT10, 16, 34, and 40) and hypo-phosphorylated early or at the end of day (e.g. CT4, 22, 28, and 46) throughout two daily cycles (Fig. 2F) indicating the strong circadian rhythm of cultured astrocytes. Importantly, the levels of HERP proteins oscillated peaking at midday (CT10, CT34) and reaching a minimum early in the day or late at night (CT4, CT22, CT36), a pattern similar to that of its mRNA (Fig. 2F and 2G). Finally, to investigate whether the rhythmic expression of Herp is regulated by a circadian clock, we examined Herp mRNA and protein expression patterns in primary astrocyte cultured from Bmal1-/- mice. In Bmal1-/- astrocyte cultures, the rhythmic expression patterns of Per2, Rev-Erbα, and Herp were abrogated and their expression was maintained at trough levels (Fig. 2A). HERP protein levels were constant throughout the daily cycles (Fig. 2H and 2I). Collectively, these observations indicated that the expression of Herp exhibits a robust circadian rhythm that is controlled by BMAL1 in cultured mouse astrocytes. These results further suggest that HERP controlled cellular processes would vary according to CT.

Herp is rhythmically expressed in mouse cultured astrocytes and its expression is controlled by BMAL1.

(A, B) Select core clock genes exhibiting rhythmic expression in cultured astrocytes. Circadian clock-synchronized cultured astrocytes from WT (black) and Bmal1-/- (orange) mice were harvested at the indicated time (CT0 = TASS 8 hour). Expression was analyzed for rhythmicity using MetaCycle, and expression levels of given genes were quantified using RNA-seq (A) and real-time qRT-PCR (B) data. Values are means ± SEM (n = 2; p-values are indicated by insets in graphs). White and gray backgrounds represent subjective day and subjective night, respectively. (C - E) Assessment of Herp siRNA for use in characterizing anti-HERP antisera. Cultured astrocytes were transfected with the indicated siRNA (20nM) and processed for real-time qRT-PCR (C) and Western blot (D, E) analyses at 48 hours post transfection. (C) siRNA-mediated knockdown of Herp mRNA. Values are means ± SEM (n = 2; *p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; independent t-test). (D) Representative Western blot image of HERP knockdown (from three independent experiments). GAPDH served as loading control. (E) Densitometric quantification of HERP levels, normalized to GAPDH levels. Values are means ± SEM (n = 3; ***p < 0.0005; Student’s t-test). (F–I) BMAL1-dependence of HERP expression. Circadian clock-synchronized cultured astrocytes from WT (F, G) and Bmal1-/- (H, I) mice were harvested at the indicated times (CT0 = TASS 8 hours) and processed for Western blot analysis. HERP levels were normalized to total ERK (tERK), which served as a loading control. tERK values at different times were normalized to those at CT4 (defined as 1). P-values are indicated by insets in graphs. White and gray backgrounds represent subjective day and subjective night, respectively. (F, G) BMAL1 and HERP expression over time in astrocytes from WT mice. (F) Representative Western blot image (from five independent experiments). (G) Densitometric quantification of Western blot data from WT mice. (H, I) BMAL1 and HERP expression over time in astrocytes from Bmal1-/- mice. (H) Representative Western blot image (from two independent experiments). (I) Densitometric quantification of Western blot data from Bmal1-/- mice.

Herp knockdown alters ATP-induced ER Ca2+ release

Next, we sought to investigate whether HERP-controlled processes are under circadian regulation. We examined HERP regulated ER Ca2+ response in the astrocytes, which have previously been reported in other cell types (Paredes et al., 2016; Torrealba et al., 2017) employing the organelle-specific fluorescent Ca2+ indicator G-CEPIA1er(Suzuki et al., 2014) (ER specific), R-GECO1(Wu et al., 2013) (cytosol specific), and mito-R-GECO1(Zhao et al., 2011) (mitochondria specific). ATP serves not only as a fundamental energy source but also functions as an active intercellular messenger. In astrocytes, application of ATP induces an increase in intracellular Ca2+ (Neary et al., 1988), a process mediated by ATP binding to the Gq/G11 coupled P2Y receptor, which acts through activation of PLC-β and subsequent hydrolysis of PIP2 to produce diacylglycerol (DAG) and IP3, the latter of which then mobilize Ca2+ through IP3Rs on the ER membrane (Baryshnikov et al., 2003; James & Butt, 2001). We treated control and Herp siRNA treated astrocytes with 100μM ATP and monitored subcellular changes in Ca2+. ATP treatment rapidly decreased ER Ca2+ in control astrocytes and this response was greater in Herp siRNA treated astrocytes (Fig. 3A - 3C). The ER as the main Ca2+ store in the cell releasesCa2+ which is then rapidly transmitted to other organelles such as mitochondria and lysosomes and altered cytosolic Ca2+ signals. Although ATP treatment did not significantly alter cytosolic Ca2+ signal in control astrocytes, it greatly increased cytosolic Ca2+ signals in Herp-knockdown astrocytes (Fig. 3D – 3F). Consistently, mitochondrial Ca2+ was greatly increased under Herp siRNA condition compared with control (Fig. 3G – 3I). These results indicated that HERP controls ER Ca2+ responses in astrocytes thereby impacting diverse subcellular signaling processes. Next, we examined the mechanism by which HERP controls ER Ca2+.There are three subtypes of IP3Rs encoded by the genes Itpr1, Itpr2, and Itpr3. In our cultured astrocytes, transcriptome analysis indicated Itpr1 as the most prevalent, followed by Itpr2, while Itpr3 expression was minimal (Fig. S4). Notably, expression of Itpr subtypes was not rhythmic. We then assessed the ITPR1 and ITPR2 protein levels in control and Herp-knockdown astrocytes and found that the levels of both ITPR1 and ITPR2 were slightly but statistically significantly increased in in Herp siRNA-treated astrocytes compared with controls (Fig. 3J – 3M). This suggests that HERP regulates ER Ca2+ release via IP3R degradation in astrocytes.

Herp knockdown altered ATP-induced ER Ca2+ response.

(A-I) Imaging of subcellular ATP-induced Ca2+ signals in Herp-knockdown and control cultured astrocytes. Cultured astrocytes were co-transfected with 20 μM non-targeting (CTRL) siRNA or Herp siRNA together with ER (A–C), cytosol (D–F) or mitochondrial (Mito; G–I) compartment-specific Ca2+ indicator (denoted at left). At 48 hours post transfection, cultured astrocytes were stimulated with 100 µM ATP and Ca2+ imaging analysis was performed. Images were acquired every 3 seconds. (A, D, G) Representative time-lapse images of each Ca2+ indicator. (B, E, H) ΔF/F0 values over time following ATP application. (C, F, I) Area under the curve values, calculated from panels B, E, and H. (A–C) CTRL siRNA, n = 19; Herp siRNA, n = 22. (D–F) CTRL siRNA, n = 20; Herp siRNA, n = 25. (G–I) CTRL siRNA, n = 16; Herp siRNA, n = 16. (J–M) ITPR1 and ITPR2 protein levels in control and Herp-knockdown astrocytes. Cultured astrocytes were transfected with the indicated siRNA (20nM) and processed for Western blot analysis 48 hours after transfection. Vinculin and GAPDH served as loading control for ITPRs and HERP, respectively. (J and K) ITPR1 protein levels. Itpr1 siRNA-transfected astrocytes were included as controls. (J) Representative Western blot image (from twelve independent experiments). (K) Densitometric quantification of Western blot data showing relative levels of ITPR1 in Herp siRNA-transfected astrocytes compared with CTRL siRNA transfected astrocytes. (L, M) ITPR2 protein levels. Itpr2 siRNA-transfected astrocytes were included as controls. (L) Representative Western blot image (from five independent experiments). (M) Densitometric quantification of Western blot data showing relative levels of ITPR2 in Herp siRNA-transfected astrocytes compared with CNTL astrocytes. Values in graphs are means ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; independent t-test). See also S4.

ATP-induced ER Ca2+ release varies according to CT

Because the amount of HERP exhibited circadian variation and HERP regulates ER Ca2+ release, we next investigated whether ATP-induced ER Ca2+ responses differ different depending on CT. After synchronizing cultured astrocytes, we measured ER Ca2+ following ATP treatment at peak (CT34) and trough (CT22) HERP time points (Fig. 4A). Because the levels of HERP were lower at CT22 resulting in higher IP3R levels it is conceivable that ER Ca2+ release might be greater at CT22 than at CT34. Consistent with this, ER Ca2+ decreased to a greater extent at CT22 than at CT34 upon ATP treatment (Fig. 4B – 4D). Also, in keeping with this prediction, cytosolic Ca2+ increased to a greater extent at CT22 than at CT34 following ATP treatment (Fig. 4E - 4G). Although variations in mitochondrial Ca2+ responses exhibited a consistent circadian tendency as well, but this trend did not reach statistical significance (Fig. 4H – 4J). Finally, we assessed whether the levels of ITPR1 and ITPR2 varied according to CT. The phosphorylation status of BMAL1 served as a proxy for CT (Fig. 4K). The levels of ITPR2 showed CT-dependent changes opposite those of HERP, whereas ITPR1 levels were constant at both CTs (Fig. 4K and 4L). Because Itpr2 mRNA levels did not show a rhythmic pattern (Fig. S4), we reasoned that the CT-dependent differences in ITPR2 levels were mainly driven by HERP-mediated ITPR2 degradation. The fact that ITPR1 levels were constant further support the conclusion that circadian rhythmic ER Ca2+ responses are mainly driven by oscillations in ITPR2.

ATP-induced ER Ca2+ release varies according to CT.

(A) Schematic depiction of experimental scheme from transfection to live-cell Ca2+ imaging at different CTs. (B–J) Imaging of subcellular ATP-induced Ca2+ signals in cultured astrocytes as a function of CT. Cultured astrocytes were transfected with ER (B–D), cytosolic (E–G) or mitochondrial (Mito; H–J) compartment-specific Ca2+ indicator (denoted at left) and then their circadian rhythm was synchronized by SS. At the indicated CT, astrocytes were stimulated with 100 µM ATP and Ca2+ imaging was performed. (B, E, H) Representative time-lapse images of each Ca2+ indicator. (C, F, I) ΔF/F0 values over time following ATP application. (D, G, J) Area under the curve values, calculated from panels C, F, and I. (B–D) CT22, n = 24; CT34 n = 19. (E–G) CT22, n = 33; CT34, n = 38. (H–J) CT22, n = 50; CT34, n = 54. (K, L) CT-dependent changes in ITPR1 and ITPR2. Cells were harvested at the indicated CTs and processed for Western blot analysis. Vinculin and GAPDH served as loading control for ITPR and BMAL1, respectively. (K) Representative Western blot image (from six independent experiments). (L) Densitometric quantification of Western blot data showing relative levels of ITPR1 and ITPR2 at different CTs. Values in graphs are means ± SEMs (*p < 0.05, ****p < 0.00005; independent t test). See also S4.

Finally, we sought to confirm that ATP-induced ER Ca2+ response is controlled by the circadian clock by analyzing this response in cultured astrocytes from Bmal1-/- mice. This analysis showed no CT differences in the decrease in ER Ca2+ following ATP treatment (Fig. 5A – 5C). The absence of rhythmic ER Ca2+ responses in Bmal1-/- astrocytes was associated with unchanging levels of HERP (Fig. 5D - 5E). We also noticed that the ER Ca2+ decrease was greater in Bmal1-/- astrocytes than in control astrocytes (Fig. 4D and 5C), a difference that might be explained by the fact that HERP levels in Bmal1-/- astrocytes were comparable to trough levels observed in control astrocytes (Fig. 5E). Collectively, these results clearly indicate that the circadian clock controls rhythmic ER Ca2+ responses through rhythmic oscillations in HERP levels in cultured mouse astrocytes.

CT-dependent ER Ca2+ release is abolished in cultured astrocytes from Bmal1-/- mice.

Cultured astrocytes from Bmal1-/- mice were transfected with ER Ca2+ indicator and then their circadian rhythm was synchronized by SS. At the indicated CTs, astrocytes were stimulated with 100 μM ATP and Ca2+ imaging was performed. (A) Representative time-lapse images of ER Ca2+ indicator. (B) ΔF/F0 values over time following ATP application. (C) Area under the curve values, calculated from panel B. (A–C) CT22, n = 34; CT34, n = 43. (D, E) Cultured astrocytes from Bmal1-/- mice and WT littermates were synchronized by SS. Cells were harvested at the indicated CTs and processed for Western blot analysis. (D) Representative Western blot image (from six independent experiments). GAPDH served as a loading control. (E) Relative levels of HERP in cultured astrocytes from Bmal1-/- mice and WT littermates at different CTs. Values are means ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; repeated measures two-way ANOVA).

ATP-induced S368-phosphorylation of Cx43 and gap junctional communication shows circadian variation

Syncytial coupling achieved through gap junctions is the prominent feature of astrocytes. This coupling is critical for the homeostatic function of astrocytes through enhancing diffusion and equilibration of ions, metabolites, and signaling molecules (Langer et al., 2012; Verkhratsky & Nedergaard, 2018). In addition, the astrocytic syncytium can propagate Ca2+ waves that affect nearby and remote cells through release of “gliotransmitters”, by which astrocytes can modulate synaptic functions in a far-reaching network (Pacholko et al., 2020). Connexin (Cx) family proteins are the molecular constituents of gap junction channels (GJCs) (Giaume et al., 2021). Cx43 (also known as Gja1) and Cx30 (also known as Gjb6) are the main connexins of astrocytes (Kunzelmann et al., 1999; Nagy et al., 1999). In our cultured astrocytes, the most abundantly expressed connexin was Cx43 (Fig. S5). The phosphorylation of Cx43 at Ser368 by protein kinase C (PKC), activated in response to intracellular Ca2+, is known to decrease gap junction conductance (Enkvist & McCarthy, 1992; Nimlamool et al., 2015; Solan & Lampe, 2014). Accordingly, we reasoned that the circadian variation in ATP-driven ER Ca2+ responses might differentially impact Cx43 phosphorylation according to times of day. We first confirmed that treatment of cultured astrocytes with ATP induced rapid phosphorylation of Cx43 at Ser368 (Fig. 6A and 6B). We then examined whether phosphorylation of Cx43 at Ser368 varied depending on CT. Notably, ATP-induced ER Ca2+ release was more pronounced at CT22 than at CT34 (Fig. 3). The phosphorylation of Cx43 at Ser368 in response to ATP treatment, was also significantly higher at CT22 compared with that at CT34 (Fig. 6C and 6D).

ATP-induced pCx43(S368) levels and the conductance of gap junctions varies according to CT.

CT-dependent changes in phosphorylated Cx43 (pCx43). (A–D) Cultured astrocytes, with (C, D) or without (A, B), circadian clock synchronization were treated with 100 µM ATP and processed for Western blot analysis at the indicated times. Vinculin, GAPDH, and/or β-tubulin (TUBB) served as loading controls. The intensity of pCx43(S368) and Cx43 for each sample was normalized to that of Vinculin. (A, B) Cx43 phosphorylation in unsynchronized cultured astrocytes at different times (0, 5, 15, 30 minutes) after stimulation with ATP. (A) Representative Western blot image (from more than three independent experiments). Vinculin served as a loading control. (B) Densitometric quantification of Western blot data, showing relative pCx43(S368)/Cx43 levels. Values are normalized to those for mock-treated samples at time zero (defined as 1). (C, D) Cx43 phosphorylation in SS-synchronized cultured astrocytes at CT22 and CT34 at different times (0, 5, and 15 minutes) after stimulation with ATP. (C) Representative Western blot image (from more than three independent experiments). Vinculin, GAPDH, and β-tubulin (TUBB) served as loading controls. (D) Densitometric quantification of Western blot data, showing relative pCx43(S368)/Cx43 levels. Values are normalized to those for mock-treated samples at time zero for CT22 (defined as 1). (E, F) Changes in Cx43 phosphorylation in vivo. Mice were entrained to a 12-hour light/dark cycle followed by constant dark conditions. At the indicated CTs, the prefrontal cortex area was dissected and processed for Western blot analysis. (E) Representative Western blot image (from three independent experiments). (E) Densitometric quantification of Western blot data showing relative levels of pCx43(S368). (G-L) Circadian rhythm of cultured astrocytes from (G-I) WT and (J-L) Bmal1-/- mice were synchronized by SS. At the indicated CT, gap-FRAP was performed. (G and J) Representative time-lapse images of prebleaching, bleaching and recovery condition during gap-FRAP analysis at CT22 and CT34. (H and K) Ft/F0 values over time following photobleaching (yellow rectangle). (I and L) Recovery % values, calculated from panel H and L, respectively. (G-L) CT22, n = 15; CT34, n = 11. (J-L) CT22, n = 9; CT34, n = 9. Values in graphs are means ± SEMs (*p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; independent t-test). See also S5-S6.

Next, we examined whether circadian variations in Cx43 phosphorylation occurred in vivo. To this end, we entrained mice to a 12:12 hour light/dark (LD) cycle for two weeks and then maintained them in constant darkness (DD). Probing samples under constant darkness ensured that variations in Cx43 phosphorylation were due to the internal circadian clock and not external light stimuli. Prefrontal cortex tissues were dissected at CT58 and CT70 at third days of DD and Ser368-phosphorylated Cx43 (pCx43) was measured. Intriguingly, Cx43 phosphorylation was pronounced at subjective night, consistent with our in vitro results. Finally, we sought to assess whether gap junctional communication shows CT variation in cultured astrocytes by utilizing the fluorescence recovery after photobleaching (FRAP) method. Initially, we confirmed the role of gap junctions in intercellular communication in cultured astrocytes by observing the absence of fluorescent signal recovery after applying the gap junction channel blocker, carbenoxolone (CBX) (Fig. S6A – S6C), which was in line with previous studies (Santiquet et al., 2012). Subsequently, we investigated CT dependent variations in this communication. We observed a more rapid and pronounced recovery of fluorescence in photobleached astrocytes during subjective day (e.g., CT34) compared with subjective night (e.g., CT22) (Fig. 6G – 6I). Interestingly, this variation was absent in astrocytes cultured from Bmal1-/- mice, indicating that the day and night variations in gap junctional communication is regulated by circadian clock (Fig. 6J. – 6L).

In summary, we provide a model illustrating how circadian clock influences astrocyte function through the regulation of ER Ca2+ response. In astrocytes, the BMAL1/CLOCK-controlled oscillation of HERP regulates day and night variations in ITPR2 levels. These oscillations contribute to changes in ER Ca2+ responses, which in turn associate with distinct differences between day and night in the phosphorylation of Cx43 at Ser368 and the conductance of the gap junction channel (GJC).

Discussion

It is widely acknowledged that the circadian clock orchestrates cellular functions in a highly cell type-specific manner by governing transcription (Mure et al., 2018; Zhang et al., 2014). Although there is ample evidence attesting to the importance of the circadian clock in astrocytes (Barca-Mayo et al., 2020; Barca-Mayo et al., 2017; Griffin et al., 2019; Lananna et al., 2018; Prolo et al., 2005; Tso et al., 2017), explorations of how the circadian clock precisely modulates astrocytic functions employing astrocyte-specific circadian transcriptome analyses have been limited. Our study conducted such an analysis on primary cultured cortical astrocytes and discovered that rhythmic expression of Herp underlies day/night variation in ER Ca2+ response, showing that HERP acts by altering IP3R levels. We also found that Ca2+-activated Cx43 phosphorylation varied in accordance with the daily rhythmic ER Ca2+ response. Given that astrocytes function in a network in which Ca2+ propagates through GJCs, we propose that the time-dependent control of ER Ca2+ responses could influence the synaptic activity modulating role of astrocytes (see below).

In this study, we discovered that 2.3 % of all transcripts showed circadian rhythmic expression with astrocyte specificity. Notably, numerous genes associated with key astrocyte functions such as “regulation of systemic arterial blood pressure”, “cell migration”, “metabolic processes”, “G protein coupled receptor signaling”, and “ion transport” etc. appeared to be controlled by the circadian clock. Of particular interest, Herp exhibited rhythmic expression in phase with Per2 in primary cultured astrocytes. In Bmal1-/- astrocyte cultures, Herp showed flattened expression at its nadir level, suggesting that its transcription is directly regulated by CLOCK/BMAL1 heterodimers (Fig. 2A). Although no canonical E box was found in the upstream 6 Kb region of the Herp gene, a previous ChIP-Atlas analysis (https://chip-atlas.org)(Oki et al., 2018; Zou et al., 2022) demonstrated CLOCK binding to the Herp genomic region (Annayev et al., 2014; Hong et al., 2018). Additionally, the TTFL repressors E4BP4 (Adenovirus E4 promoter Binding Protein 4), also referred to as NFIL3 (Nuclear Factor Interleukin 3), as well as the activators TEF (Thyrotroph Embryonic Factor) and HLF (Hepatic Leukemia Factor) also bind to the Herp genomic region (Dunham et al., 2012; Liu et al., 2022; Yoshitane et al., 2019). Oscillations in Herp expression have been previously documented in various mouse tissues including the lung, heart, liver, kidney, aorta, skeletal muscle, adrenal gland, pancreatic islet and SCN (Panda et al., 2002; Rakshit et al., 2016; Zhang et al., 2014). Collectively, this wealth of evidence strongly suggests that Herp is subject to direct regulation by the circadian clock. Our study further demonstrated that Herp oscillates at both mRNA and protein levels in cultured astrocytes. The significance of HERP in modulating circadian variations in Ca2+ signaling in the afore-mentioned tissues and cells is an intriguing avenue for future research. Interestingly, a recent study employing a circadian single-cell RNA seq analysis of SCN reported that Herp displays a circadian rhythm exclusively in astrocytes and oligodendrocytes (Wen et al., 2020). Given that astrocytic Ca2+ levels in the SCN demonstrated a circadian rhythm, peaking during the subjective night and are crucial for circadian timing (Brancaccio et al., 2019; Patton et al., 2022), HERP’s circadian rhythmic expression in SCN astrocytes (Wen et al., 2020) may be integral to the mechanisms of circadian timing.

We demonstrated that HERP regulated ER Ca2+ responses by degrading IP3Rs in astrocytes, aligning with findings in other cell types (Fig. 3J – 3M). (Paredes et al., 2016; Torrealba et al., 2017). There are three subtypes of IP3Rs in astrocytes encoded by the genes Itpr1, Itpr2, and Itpr3. While all three subtypes are expressed in our cultured astrocytes, Itpr3 was expressed in negligible amounts (Fig. S4). Traditionally, ITPR2 has been regarded as a major subtype closely associated with astrocyte function (Holtzclaw et al., 2002; Sharp et al., 1999) and accumulating evidence has shown that not only ITPR2, but also ITPR1 and ITPR3 are important in astrocyte Ca2+ signaling, with each subtype eliciting distinct Ca2+ responses (Sherwood et al., 2021). When Herp was downregulated, the levels of both ITPR1 and ITPR2 increased, but only the levels of ITPR2 exhibited day/night variation (Fig. 3J – 3M, Fig. 4K – 4L). The min/max ratio of HERP expression throughout the day was 0.38 under normal conditions, but was 0.072 in Herp-KD conditions. Based on our study, we reasoned that the amplitude of HERP oscillations impacted the levels of ITPR2 under normal circumstances, while having a much lesser effect on the levels of ITPR1. Thus, we conclude that the circadian control of ER Ca2+ release in response to stimulation in cultured astrocytes is attributable to ITPR2.

Our study revealed distinct circadian variations in ER and cytosolic Ca2+ responses following ATP treatment that were not reflected in mitochondrial Ca2+ responses (Fig. 4). Upon ATP stimulation, Ca2+ is released from the ER via IP3R, and subsequently enters both the cytosol and mitochondria. In the cytosol, direct Ca2+ transfer occurs through IP3R. For mitochondrial uptake, Ca2+ first transits to the intermembrane space via IP3Rs and the voltage-dependent anion channel (VDAC) complex before entering the matrix through mitochondrial calcium uniporter (MCU) (Giorgi et al., 2018). The Mito-R-GECO1 reporter used in our studies of changes in mitochondrial Ca2+ is confined within the mitochondrial matrix and thus cannot assess Ca2+ levels in in the intermembrane space (Wu et al., 2013). Thus, we think of several possibilities to account for the absence of significant circadian variations in mitochondrial Ca2+ in response to ATP stimulation relative to cytoplasmic Ca2+ change. First, substantial changes in Ca2+ levels might occur in the intermembrane space, but the influx into the matrix via MCUs could remain constant across different CT, resulting in apparent lack of significant change. Second, the amount of VDAC or the VDAC:IP3R complex on the mitochondrial membrane might vary inversely with HERP levels, potentially maintaining a consistent flow of Ca2+ into the mitochondria irrespective of the CT. Additionally, the presence of several Ca2+ extrusion mechanisms in mitochondria (Giorgi et al., 2018) points to a more complex regulation of Ca2+ compared with that in the cytosol. Although the precise mechanism remains to be elucidated, our data indicate that the circadian variation in Ca2+ release from the ER, particularly in response to stimuli, predominantly impacts cytoplasmic signaling rather than mitochondrial signaling.

Importantly, we discovered that during the subjective night (active period for mice), ATP- stimulated ER Ca2+ release was pronounced compared with that during the subjective day (rest period for mice) (Fig. 4). This finding aligns with recent studies that emphasized the role of astrocytic Ca2+ in sleep regulation. In mice, stimulus-evoked Ca2+ activity in astrocytes was found to be heightened during wakefulness and diminished as sleep progressed (Bojarskaite et al., 2020; Ingiosi et al., 2020). In flies, Ca2+ activity in astrocytes is correlated with the duration of wakefulness, suggesting that astrocytic Ca2+ is a key factor in the homeostatic regulation of sleep (Blum et al., 2021). Collectively, these reports collectively underscore the role of astrocytic Ca2+ signaling in sleep homeostasis. Our observations that ATP-stimulated ER Ca2+ release is elevated during the subjective night compared with the subjective day provides a mechanistic explanation for the differential Ca2+ levels observed between wake and sleep states. Notably, the circadian regulation of ITPR2 by HERP might be a fundamental factor in this process.

We also observed a marked increase in ATP-stimulated Cx43 (S368) phosphorylation during the subjective night compared with the subjective day (Fig. 6C and 6D.). This observation was further supported by in vivo experiments, which demonstrated higher levels of pCx43(S368) in the frontal cortex during the subjective night than during the day (Fig. 6E and 6F). What are the implications of this day/night variation in Cx43 (S368) phosphorylation? Cx43 and Cx30 are the main connexins in astrocytes, where they constitute key components of gap junctions (Kunzelmann et al., 1999; Nagy et al., 1999). Gap junctions mediate intercellular communication by providing ultrastructural cytoplasmic continuity and are thus integral to formation of the functional syncytium of astrocytes (Giaume et al., 2010; Kiyoshi & Zhou, 2019). In our study, Cx43 was the main connexin and was most abundantly expressed, whereas Cx30 expression was almost negligible (Fig. S5). Cx43 possesses a distinctive structural composition consisting of an extensive C-terminal tail, an N-terminal domain, and multiple transmembrane segments. The C-terminal tail contains sites for post-translational modifications. Specifically, Cx43 has more than 10 phosphorylation sites, and its phosphorylation is key to the control of its function (Pogoda et al., 2016; Solan & Lampe, 2014). Among these sites Ser368 is a target for phosphorylation by PKC, which becomes activated in response to elevated Ca2+ levels; notably this phosphorylation event leads to the closure of Cx43 gap junction channels (Enkvist & McCarthy, 1992; Nimlamool et al., 2015; Pogoda et al., 2016; Pun et al., 2022; Solan & Lampe, 2014, 2016). We reasoned that the circadian variation in Cx43 phosphorylation could be significant in terms of astrocyte functionality within the syncytium. Indeed, in our cultured astrocytes, gap junctional communication exhibited day and night variation (Fig.6G – 6I). Astrocytes influence synaptic activity through the release of gliotransmitters such as glutamate, GABA, D-serine, and ATP, triggered by increases in intracellular Ca2+ in response to the activity of adjacent neurons and astrocytes (Verkhratsky & Nedergaard, 2018). Importantly, this increase in Ca2+ spreads to adjacent astrocytes through GJCs (Fujii et al., 2017), influencing a large area of the neuronal network. Considering that Cx43 Ser368 phosphorylation occurs to uncouple specific pathways in the astrocytic syncytium to focus local responses (Enkvist & McCarthy, 1992), our findings suggest that astrocytes better equipped for localized responses when presented with a stimulus during the active phase in mice. Conversely, during the rest period, characterized by more synchronous neuronal activity across broad brain areas (Vyazovskiy et al., 2009) higher GJC conductance might allow astrocytes to exert control over a larger area. In support of this idea, recent study showed that synchronized astrocytic Ca2+ activity advances the slow wave activity (SWA) of the brain, a key feature of non-REM sleep (Szabó et al., 2017). Blocking GJC was found to reduce SWA, further supporting this interpretation. However, conflicting findings have also been reported. For instance, Ingiosi et al. (Ingiosi et al., 2020) found that astrocytic synchrony was higher during wakefulness than sleep in the mouse frontal cortex. Whether these differing results in astrocyte synchrony during resting and active periods are attributable to differences in experimental context (e.g., brain regions, sleep-inducing condition) remains unclear. Indeed, astrocyte Ca2+ dynamics during wakefulness/sleep vary according to brain regions (Tsunematsu et al., 2021). While the extent of astrocyte synchrony might differ depending on brain region and/or stimulus, on our results suggest the conclusion that that the baseline state of astrocyte synchrony, which is affected by GJC conductance, varies with the day/night cycle.

The presence of various glial subpopulations, each with high expression of multiple connexins, contributes to the formation of a “panglial syncytium.” This allows gap junction channels to form between astrocytes and other glial or non-glial cells (Fróes et al., 1999; Rash et al., 1997). Astrocytic GJCs connecting both oligodendrocytes and ependymal cells lining the ventricles have been reported (Rash et al., 1997). Recent studies in mice with cell type-specific deletion of connexins have demonstrated that removal of the astrocytic connexins (Cx30 and Cx47) results in the total disruption of panglial coupling. Moreover, this coupling can be maintained if at least one allele of either connexin is retained (Griemsmann et al., 2015). Based on our experimental findings, we conclude that phosphorylation of Cx43 at Ser368 in response to a stimulus is contingent on CT and consequently influences the permeability of GJCs in a CT dependent manner. Drawing from previous research, this could have substantial implications not only for GJs between astrocytes, but also for those linking astrocytes with other cell types, such as oligodendrocytes, and other cell types.

In conclusion, our findings significantly enhance the understanding of the circadian regulation of astrocytic functions, particularly through the rhythmic expression of HERP and its impact on ER Ca²⁺ signaling and Cx43 phosphorylation. This study underscores the intricate interplay between the circadian clock and astrocyte physiology, highlighting the pivotal role of HERP in modulating key astrocytic processes such as sleep, gap junction communication and synaptic modulation. The day/night variations in ER Ca²⁺ response and Cx43 phosphorylation, and their implications for astrocytic network dynamics and neuronal interactions, open new avenues for exploring the broader implications of astrocytic circadian rhythms in brain function and health. Thus, this work not only contributes to our fundamental understanding of astrocyte biology, it also has potential implications for addressing neurological disorders where circadian and astrocytic dysfunctions intersect.

Materials and Methods

Animals

All procedures for the care and use of laboratory animals were approved by the Institutional Animal Care and Use Committee (IACUC) of Ajou University School of medicine. Initially, Bmal1-/+ mice (B6.129-Bmal1tm1Bra/J; Jackson Laboratory), #009100)(Bunger et al., 2000) were purchased from Jackson Laboratory(USA) and housed in a specific pathogen-free environment at the Animal Research Center of Ajou University Medical Center, with a standard 12-hour light/dark cycle and free access to food. In experiments employing primary cultured astrocytes from Bmal1-/- mice, wild type (WT) littermates were used as controls. Primary cultured astrocytes involving only WT mice used 1 day-old C57BL/6 mice (Koatch Inc., Korea) For animal studies, 7-8 week-old adult C57BL/6 male mice were first entrained in a standard 12-hour light/dark cycle for 2 weeks and then maintained in a constant darkness (DD). On the third day of DD, mice were sacrificed at indicated circadian time (CT) and its prefrontal cortex (PFC) was dissected.

Primary astrocyte culture

Cultures of primary astrocytes were prepared according to a previous report (Choi et al., 2018) with minor modifications. In brief, 1 day old pups were anesthetized by hypothermia and their cerebral cortices were dissected out and triturated in Modified Eagle’s medium (MEM; Welgene(Korea), LM 007-11) containing 10% fetal bovine serum (FBS; Hyclone(USA), SV30207.02), penicillin/streptomycin (Gibco(USA), 15140-122), 10 µM HEPES (Gibco, 15630080) and GlutaMAX (Gibco, 35050-061) yielding a single-cell suspension. Cells were plated into 75 cm2 T-flasks (1pup/flask) and incubated at 37°C in a humidified 5% CO2 incubator for 2 weeks. Primary astrocytes were then incubated in serum-free MEM for 7 days, after which astrocytes were detached from the T-flask with 0.25% trypsin and plated on culture dishes (1.0X106cells/60 mm dish) for further experiments.

Circadian transcriptome analysis

The experimental scheme for synchronizing circadian clocks of cultured astrocyte is depicted in Figure 1A. In brief, astrocytes were grown in dishes (1.0 X 106cells/60 mm dish) until reaching complete confluence and then incubated in serum-free MEM for 72 hours. Cells were then subjected to serum shock (SS), a well-established procedure for synchronizing the circadian clock of cultured cells (Balsalobre et al., 1998), by exchanging the medium for MEM containing 50% horse serum. Twelve hours after SS, astrocytes were harvested at 4-hour intervals for 2 days for RNA sequencing (RNA-seq).

Total RNA was extracted and purified from harvested astrocytes using an RNeasy Plus Micro Kit (Qiagen (Germany), 74034). RNA quality and quantity assessments, RNA-Seq library construction, and next-generation sequencing analysis were performed by Macrogen Inc. (Korea). Total RNA integrity and library size were analyzed using an Agilent Technologies 2100 Bioanalyzer. Paired-end raw reads, generated using an Illumina HiSeq 2000 system, were aligned to the mm10 mouse genome using STAR aligner (v.2.5.2b)(Dobin et al., 2013). Gene expression was subsequently quantified by calculating transcripts per million (TPM) using Ensemble gene (release 82) annotations. The threshold for defining expressed transcripts was set using a Gaussian mixture model in the R package ‘mixtools’

To detect circadian oscillating transcripts, we employed two circadian oscillation detection methods BioCycle(Agostinelli et al., 2016) and MetaCycle (Wu et al., 2016). BioCycle contains various types of oscillating signal patterns, and deep learning methods was applied to develop and train algorithms. MetaCycle incorporates ARSER, JTK_CYCLE, and Lomb-Scargle algorithm for period signal detection (Wu et al., 2016). We tested both methods and decided to use union of rhythmic transcript lists from two methods. For the biological and functional annotations for the identified circadian clock transcripts, we used Metascape (https://metascape.org)(Zhou et al., 2019) with default express analysis option.

Quantitative reverse transcription polymerase chain reaction (qRT-PCR)

Total RNA was extracted from cells and purified using the RNeasy Plus Micro Kit (Qiagen, 74034). 1μg of total RNA was reverse transcribed using an oligo-dT primer and PrimeScript RTase (TaKaRa (Japan), 2680A). Quantitative real-time PCR was performed using a Rotor Gene Q (Qiagen) with TB Green Premix Ex Taq (Takara, RR420A) The specific primers used were provided in the Supplementary Table 1. Noncycling mRNA encoding HPRT was used to normalize gene expression. The data were analyzed using Rotor Gene 6000 software, and the relative mRNA levels were quantified using the 2-2-ΔΔCt method in which ΔΔCt=[(Ct target – Ct HPRT) of experimental group] – [(Ct target – Ct HPRT) of control group].

siRNA transfection and Immunoblotting

The following On-Target plus SMARTpool siRNAs were purchased from Dharmacon (USA) : Non-targeting control siRNA(D-001810-01-50), Herp siRNA (L-049714-01-0005), Itpr1 siRNA(L-040933-00-0005), and Itpr2 siRNA(L-041018-00-0005). After cells had reached greater than 90% confluence, the medium was changed to serum-free MEM and cells were incubated for 72 hours. Cells were then transfected with siRNA using Lipofectamine RNAiMAX (ThermoFisher (USA), 13778150) according to manufacturer’s protocol. Experiments were performed 48 hours after transfection.

For immunoblotting, astrocytes were lysed using modified-RIPA buffer (50 mM Tris-HCl pH 7.4, 1% NP-40, 0.5% sodium deoxycholate, 150 mM NaCl) supplemented with protease inhibitor cocktail (Sigma-Aldrich (USA), P8340) and phosphatase inhibitor cocktails 2 and 3 (Sigma-Aldrich, P5726 and P0044). Proteins were separated by sodium dodecyl sulfate-polyacrylamide gel electrophoresis (SDS-PAGE) and transferred to polyvinylidene fluoride membranes. After blocking with 5% skim milk, membranes were incubated overnight at 4°C with the following primary antibodies at the indicated dilutions: anti-BMAL1 (Abcam (UK), ab93806), 1:2000; anti-HERP (Abcam, ab150424), 1:1000; anti-ITPR1 (Abcam, ab264284), 1:1000; anti-ITPR2 (Alomone Labs (Israel), ACC-116),1:1000; anti-Cx43 (Sigma, C6219), 1:5000; anti-pCx43 (Ser368) (CST (USA), 3511), 1:1000; anti-GAPDH (Novus (USA), NB100-56875), 1:5000; anti-vinculin (Sigma-Aldrich, V4505), 1:5000; and anti-total ERK (CST, 9102), 1:5000. Membranes were washed with Tris-buffered saline containing 0.5% Tween 20 (TBST), incubated with secondary antibodies, and visualized using an enhanced chemiluminescence system. Protein levels were quantified densitometrically by measuring band intensities using Image J software.

Ca2+ reporter plasmid transfection and imaging analysis

After cells reach greater than 90% confluence, the medium was changed to serum free MEM. Ca2+ measurement was conducted using pCMV-G-CEPIA1er, a gift from Masamitsu Iino (Addgene (USA) plasmid #58215)(Suzuki et al., 2014), and CMV-R-GECO1 and CMV-mito-R-GECO1, gifts from Robert Campbell (Addgene plasmids #32444, #46021)(Wu et al., 2013; Zhao et al., 2011). These plasmids, encoding Ca2+ indicators, were transfected into cells 72 hours after switching to serum-free conditions, with or without siRNA, using Lipofectamine 2000 according to manufacturer’s protocol. Forty-eight hours after transfection, the medium was replaced with Hanks balanced salt solution (HBSS) without Ca2+ and Mg2+ (Gibco, 14175095), and fluorescence images were acquired every 3 seconds using a Nikon A1R Confocal Microscope system equipped with a x60 1.4 NA Plan-Apochromat objective (Nikon Instrument, Inc. (Japan)) at 37°C in a humidified 5% CO2 atmosphere. These experiments were performed at the Three-Dimensional Immune System Imaging Core Facility of Ajou University (Korea). Images were analyzed using an NIS elements C program. Fluorescence intensity was measured at each time point (Ft) and ΔF values were calculated from (Ft – F0). F0 values, used for normalization, were defined by averaging 10 frames before stimulation.

Gap-FRAP (fluorescence recovery after photobleaching) analysis

Astrocyte communication was assessed using the gap-FRAP assay with slight modifications from a previous report (Santiquet et al., 2012). This involved quantifying the transfer of the fluorescent dye calcein from cells to another cell that had been subjected to laser bleaching to remove calcein fluorescence. Thirty minutes prior to imaging, cells were loaded with 0.5uM Calcein-AM (Invitrogen (USA), C1430) and 0.05% Pluronic-127 (Invitrogen, P6866) at 37°C. A 30min incubation period allowed for the dye to diffused and be cleaved into fluorescent calcein, after which cells were rinsed twice with serum free MEM. In experiments involving CBX treatment, CBX (20μM) was applied for an additional 30 minutes after calcein incubation. Immediately before imaging, the medium was switched to HBSS without Ca2+ and Mg2+ (Gibco, 14175095). Utilizing a Nikon A1R Confocal Microscope system equipped with a 488nm argon laser and a x60 1.4 NA Plan-Apochromat objective (Nikon Instruments, Inc.), photobleaching was performed on ROI cell, and fluorescence was measured under 37°C in a humidified atmosphere of 5% CO2. Fluorescence intensity was captured in a confocal plane during the following intervals: five pre-bleach images, a 15-second 100% power laser pulse at a stimulation speed of 5.3 lines/sec in the ROI cell, and an image taken every 3 seconds for a 3-minute post-bleach period. Images were analyzed using NIS elements C program. Fluorescence intensity was measured at each time point (Ft) and normalized intensity values were calculated as Ft/F0. F0 values, used for normalization, were defined by averaging 5 frames before bleaching. Recovery % calculated as (FFR-FA)/(F0-FA)x100, where FFR is full recovery fluorescence intensity after photobleaching and FA is the fluorescence intensity right after photobleaching.

Quantification and statistical analysis

Statistical analyses were performed using GraphPad Prism 8 software. Differences between two groups were analyzed using independent t-tests. Differences among three or more groups were analyzed by one-way analysis of variance (ANOVA) with Tukey’s post hoc test. A repeated measures two-way ANOVA was employed for comparison of multiple groups and conditions. Differences were considered significant for p-values < 0.05.

Data and code availability

Circadian RNA sequencing data have been deposited at GEO under accession number GEO: GSE254678.

Any additional information required to reanalyze the data reported in this paper is available from the E.Y.K upon request.

eLife Assessment

This is a useful study that identifies circadian changes in the gene expression profile of cultured mouse astrocytes. Mechanistic details linking circadian rhythmicity in HERP, a regulator of calcium signals in the endoplasmic reticulum, to altered phosphorylation of Connexin 43 remain currently incomplete. With improved manuscript clarity and statistical analysis, this work could be of interest to the field of astrocyte and circadian biology.

Reviewer #1 (Public Review):

Summary:

In Ryu et al., the authors use a cortical mouse astrocyte culture system to address the functional contribution of astrocytes to circadian rhythms in the brain. The authors' starting point is transcriptional output from serum-shocked culture, comparative informatics with existing tools and existing datasets. After fairly routine pathway analyses, they focus on the calcium homeostasis machinery and one gene, Herp, in particular. They argue that Herp is rhythmic at both mRNA and protein levels in astrocytes. They then use a calcium reporter targeted to the ER, mitochondria, or cytosol and show that Herp modulates calcium signaling as a function of circadian time. They argue that this occurs through the regulation of inositol receptors. They claim that the signaling pathway is clock-controlled by a limited examination of Bmal1 knockout astrocytes. Finally, they switch to calcium-mediated phosphorylation of the gap junction protein Connexin 43 but do not directly connect HERP-mediated circadian signaling to these observations. While these experiments address very important questions related to the critical role of astrocytes in regulating circadian signaling, the mechanistic arguments for HERP function, its role in circadian signaling through inositol receptors, the connection to gap junctions, and ultimately, the functional relevance of these findings is only partially substantiated by experimental evidence.

Strengths:

- The paper provides useful datasets of astrocyte gene expression in circadian time.

- Identifies HERP as a rhythmic output of the circadian clock.

- Demonstrates the circadian-specific sensitivity of ATP -> calcium signaling.

- Identifies possible rhythms in both Connexin 43 phosphorylation and rhythmic movement of calcium between cells.

Weaknesses:

- It is not immediately clear why the authors chose to focus on Ca2+ homeostasis or Herp from their initial screens as neither were the "most rhythmic" pathways in their primary analyses.

- It would have been interesting (and potentially important) to know whether various methods of cellular synchronization would also render HERP rhythmic (e.g., temperature, forskolin, etc). If Herp is indeed relatively astrocyte-specific and rhythmic, it should be easy to assess its rhythmicity in vivo.

- The authors show that Herp suppression reduces ATP-mediated suppression of calcium whereas it initially increases Ca2+ in the cytosol and mitochondria and then suppresses it. The dynamics of the mitochondrial and cytosolic responses are not discussed in any detail and it is unclear what their direct relationship is to Herp-mediated ER signaling. What is the explanation for Herp (which is thought to be ER-specific) to calcium signaling in other organelles?

- What is the functional significance of promoting ATP-mediated suppression of calcium in ER?

- The authors then nicely show that the effect of ATP is dependent on intrinsic circadian timing but do not explain why these effects are antiphase in cytosol or mitochondria. Moreover, the ∆F/F for calcium in mitochondria and cytosol both rise, cross the abscissa, and then diminish - strongly suggesting a biphasic signaling event. Therefore, one wonders whether measuring the area under the curve is the most functionally relevant measurement of the change.

- Why are mitochondrial and cytosolic calcium not also demonstrated for Bmal1 KO astrocytes?

- The authors claim that Herp acts by regulating the degradation of ITPRs but this hypothesis - rather central to the mechanisms proposed in this study - is not experimentally substantiated.

- There is no clear demonstration of the functional relevance of the circadian rhythms of ATP-mediated calcium signaling.

Reviewer #2 (Public Review):

Summary:

The article entitled "Circadian regulation of endoplasmic reticulum calcium response in mouse cultured astrocytes" submitted by Ryu and colleagues describes the circadian control of astrocytic intracellular calcium levels in vitro.

Strengths:

The authors used a variety of technical approaches that are appropriate

Weaknesses:

Statistical analysis is poor and could lead to a misinterpretation of the data

Several conceptual issues have been identified.

Overinterpretation of the data should be avoided. This is a mechanistic paper done completely in vitro, all references to the in vivo situation are speculative and should be avoided.

Reviewer #3 (Public Review):

Astrocyte biology is an active area of research and this study is timely and adds to a growing body of literature in the field. The RNA-seq, Herp expression, and Ca2+ release data across wild-type, Bmal1 knockout, and Herp knockdown cellular models are robust and lend considerable support to the study's conclusions, highlighting their importance. Despite these strengths, the manuscript presents a gap in elucidating the dynamics of HERP and the involvement of ITPR1/2 in modulating Ca2+ release patterns and their circadian variations, which remains insufficiently supported and characterized. While the Connexin data underscore the importance of rhythmic Ca2+ release triggered by ATP, the relationship here appears correlational and the role of HERP and ITPR in Cx function remains to be characterized. Moreover, enhancing the manuscript's clarity and readability could significantly benefit the presentation and comprehension of the findings.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

In Ryu et al., the authors use a cortical mouse astrocyte culture system to address the functional contribution of astrocytes to circadian rhythms in the brain. The authors' starting point is transcriptional output from serum-shocked culture, comparative informatics with existing tools and existing datasets. After fairly routine pathway analyses, they focus on the calcium homeostasis machinery and one gene, Herp, in particular. They argue that Herp is rhythmic at both mRNA and protein levels in astrocytes. They then use a calcium reporter targeted to the ER, mitochondria, or cytosol and show that Herp modulates calcium signaling as a function of circadian time. They argue that this occurs through the regulation of inositol receptors. They claim that the signaling pathway is clock-controlled by a limited examination of Bmal1 knockout astrocytes. Finally, they switch to calcium-mediated phosphorylation of the gap junction protein Connexin 43 but do not directly connect HERP-mediated circadian signaling to these observations. While these experiments address very important questions related to the critical role of astrocytes in regulating circadian signaling, the mechanistic arguments for HERP function, its role in circadian signaling through inositol receptors, the connection to gap junctions, and ultimately, the functional relevance of these findings is only partially substantiated by experimental evidence.

Strengths:

- The paper provides useful datasets of astrocyte gene expression in circadian time.

- Identifies HERP as a rhythmic output of the circadian clock.

- Demonstrates the circadian-specific sensitivity of ATP -> calcium signaling.

- Identifies possible rhythms in both Connexin 43 phosphorylation and rhythmic movement of calcium between cells.

Weaknesses:

- It is not immediately clear why the authors chose to focus on Ca2+ homeostasis or Herp from their initial screens as neither were the "most rhythmic" pathways in their primary analyses.

We appreciate the reviewer’s comment. We chose to focus on Ca2+ homeostasis processes because intracellular Ca2+ signaling plays crucial role in numerous astrocyte functions and is notably associated with sleep/wake status of animals, which is our primary interest (Bojarskaite et al., 2020; Ingiosi et al., 2020; Blum et al., 2021; Szabó et al., 2017). Among the genes involved in calcium ion homeostasis, Herp exhibited the most robust rhythmicity (supplementary table 1). The rationale for our focus on Ca2+ homeostasis and Herp is explained in the results section (line 143-150). We hope this provides a clear justification for our focus.

- It would have been interesting (and potentially important) to know whether various methods of cellular synchronization would also render HERP rhythmic (e.g., temperature, forskolin, etc). If Herp is indeed relatively astrocyte-specific and rhythmic, it should be easy to assess its rhythmicity in vivo.

Thank you for the reviewer’s insightful comment. In response, we examined HERP expression in cultured astrocytes synchronized using either Dexamethasone or Forskolin treatment. We found that Herp exhibited rhythmic expression at both the the mRNA and protein levels under these conditions. These results have been added to Figure S3 and are explained in the manuscript (lines 173-175).

Additionally, we measured HERP levels in the prefrontal cortex of mice at CT58 and CT70 and found no rhythmicity, as shown in Author response image 1. Given that Herp is expressed in various brain cell types, including microglia, endothelial cells, neurons, oligodendrocytes, and the astrocytes- with the highest expression in microglia(Cahoy et al., 2008), we reason that the potential rhythmic expression of HERP in astrocytes might be masked by its continuous expression in other cell types. Nonetheless, to assess HERP rhythmicity specifically in astrocytes in vivo, we attempted immunostaining using several anti-HERP antibodies, but none were successful. Consequently, we were unable to determine whether HERP exhibits rhythmic expression in astrocytes in vivo.

HERP levels were constant at CT58 and CT70.

(A, B) Mice were entrained under 12h:12h LD cycle and maintained in constant dark. Prefrontal cortices were harvested at indicated time and processed for Western blot analysis. Representative image shows three independent samples. (B) Quantification of HERP levels normalized to VINCULIN. Values in graphs are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; t-test)

- The authors show that Herp suppression reduces ATP-mediated suppression of calcium whereas it initially increases Ca2+ in the cytosol and mitochondria and then suppresses it. The dynamics of the mitochondrial and cytosolic responses are not discussed in any detail and it is unclear what their direct relationship is to Herp-mediated ER signaling. What is the explanation for Herp (which is thought to be ER-specific) to calcium signaling in other organelles?

Our examination of cytosolic and mitochondrial Ca2+ responses was aimed at corroborating HERP’s effect on ER Ca2+ response. Upon ATP stimulation, Ca2+ is released from the ER via IP3R receptors (IP3Rs) and subsequently transmitted to other organelles including mitochondria (Carreras-Sureda et al., 2018; Giorgi et al., 2018). Ca2+ is directly transferred to the cytosol by IP3Rs located on the ER membrane, and to the mitochondria through a complex formed by IP3R and the voltage-dependent anion channel (VDAC) on the mitochondria (Giorgi et al., 2018). Consistent with previous reports, we observed an increase of cytosolic and mitochondrial Ca2+ levels accompanied by decrease in ER Ca2+ levels following ATP treatment (See Fig. 3B, E, H, control siRNA). The ATP-stimulated ER Ca2+ release was enhanced by Herp knockdown. We reasoned that if Ca2+ release was enhanced, then cytosolic and mitochondrial Ca2+ uptakes would also be enhanced. The results were consistent with our hypothesis (See Fig. 3B, E, H, Herp siRNA). These observations are described in the Results section (lines 202-208) and in the Discussion (lines 333-348). We hope this explanation clarifies the relationship between Herp-mediated ER Ca2+ response and Ca2+ response in other organelles. Thank you for your consideration.

- What is the functional significance of promoting ATP-mediated suppression of calcium in ER?

In astrocytes, intracellular Ca2+ plays crucial role in regulating several processes. In this study, among various downstream effects of intracellular Ca2+, we examined the gap junction channel (GJC) conductance, which affects astrocytic communication. As discussed in the manuscript (lines 357-381), circadian variation in HERP results in rhythmic Cx43 (S368) phosphorylation linked with GJC conductance. We propose that during the subjective night phase, heightened ATP induced ER Ca2+ release reduces GJC conductance, uncoupling astrocytes from the syncytium, making them better equipped for localized response. On the other hand, during the subjective day phase, increased GJC conductance may allow astrocytes to control a larger area for synchronous neuronal activity which is a key feature of sleep.

- The authors then nicely show that the effect of ATP is dependent on intrinsic circadian timing but do not explain why these effects are antiphase in cytosol or mitochondria.

Moreover, the ∆F/F for calcium in mitochondria and cytosol both rise, cross the abscissa, and then diminish - strongly suggesting a biphasic signaling event. Therefore, one wonders whether measuring the area under the curve is the most functionally relevant measurement of the change.

We appreciate the reviewer’s insightful comments. As explained in our previous response, Ca2+ released from the ER is transferred to the cytosol and mitochondria. This transfer explains why the fluorescent intensities of cytosolic and mitochondrial Ca2+ indicators show anti-phasic responses to those of the ER.

We agree that cytosolic and mitochondrial Ca2+ responses may be biphasic. The decrease below the abscissa in mitochondria and cytosol likely reflects Ca2+ extrusion from these organelles. However, our primary focus was on the initial uptake of Ca2+ following ER Ca2+ release. Thus, when calculating the area under the curve (AUC), we measured the area between the ∆F/F graph and the y=0 (X-axis) for both mitochondria and cytosol. We reason that the measuring the area under the curve (above the abscissa) fits with our objective.

While addressing your concerns, we noticed errors in the Y-axis labels of Fig. 3C, 4D, and 5C. For the ER Ca2+ dynamics, we measured the area above curve. These mistakes have now been corrected.

- Why are mitochondrial and cytosolic calcium not also demonstrated for Bmal1 KO astrocytes?

In two sets of experiments (Fig. 3 and Fig. 4), we demonstrated that the increase in cytosolic and mitochondrial Ca2+ aligns with ER Ca2+ release. Since there were no circadian time differences in ER Ca2+ release in the Bmal1 KO cultures, we concluded that it was unnecessary to measure Ca2+ levels in the mitochondria and cytosol. Additionally, our primary focus is on the ER Ca2+ response rather than the Ca2+ dynamics in subcellular organelles. We hope this clarifies our rationale and maintains the focus of our study.

- The authors claim that Herp acts by regulating the degradation of ITPRs but this hypothesis - rather central to the mechanisms proposed in this study - is not experimentally substantiated.

We appreciate the reviewer’s insightful comments regarding the role of HERP in the degradation of IP3Rs. In the original manuscript, we demonstrated that treating cells with Herp siRNA leads to an increase in the levels of ITPR1 and ITPR2, suggesting that HERP might be involved in the regulation of IP3Rs stability. This observation is consistent with previous studies, which showed that Herp siRNA treatment increases ITPR levels in HeLa and cardiac cells (Paredes et al., 2016; Torrealba et al., 2017). Torrealba et al. also showed that HERP regulates the polyubiquitination of IP3Rs. Based on our results and previous reports, we hypothesized that HERP similarly regulates ITPR degradation in cultured astrocytes.

However, as the reviewer rightly pointed out, further evidence is needed to confirm that HERP specifically regulates ITPR degradation. To address this, we conducted new experiments examining the effect of XesC, an inhibitor of IP3Rs, on ER Ca2+ release. The treatment of XesC reduced the ER Ca2+ release and abolished the enhancement of ER Ca2+ release by Herp KD. These results demonstrated that HERP influences ER Ca2+ response through IP3Rs. These new findings have been added to Fig. 3N – 3P and explained in the Results section (lines 217-221).

We believe these additional experiments and clarifications strengthen our hypothesis that HERP regulates IP3R degradation, thereby modulating ER Ca2+ responses.

- There is no clear demonstration of the functional relevance of the circadian rhythms of ATP-mediated calcium signaling.

As mentioned in the previous response, we examined Cx43 phosphorylation linked with GJC conductance in the context of ATP-mediated Ca2+ signaling. Our results demonstrated circadian variations in Cx43 Ser368 phosphorylation leading to variations of gap junction channel (GJC) conductance (Fig. 6C – F and Fig. 7D - I). We have discussed the significance of this circadian rhythm in ATP driven ER Ca2+ signaling concerning astrocytic function during sleep/wake states in the manuscript (lines 357 – 382) as follows.

“ATP-stimulated Cx43 (S368) phosphorylation is higher at 30hr (subjective night phase) than at 42hr (subjective day phase) (Fig. 6C and 6D.), a finding further supported by in vivo experiments showing higher pCx43(S368) levels in the prefrontal cortex during the subjective night than during the day (Fig. 6E and 6F). What are the implications of this day/night variation in Cx43 (S368) phosphorylation? We reasoned that the circadian variation in Cx43 phosphorylation could significantly impact astrocyte functionality within the syncytium. Indeed, our cultured astrocytes exhibited circadian phase-dependent variation in gap junctional communication (Fig.7D – 7F). Astrocytes influence synaptic activity through the release of gliotransmitters such as glutamate, GABA, D-serine, and ATP, triggered by increases in intracellular Ca2+ in response to the activity of adjacent neurons and astrocytes (Verkhratsky & Nedergaard, 2018). Importantly, this increase in Ca2+ spreads to adjacent astrocytes through GJCs (Fujii et al., 2017), influencing a large area of the neuronal network. Considering that Cx43 Ser368 phosphorylation occurs to uncouple specific pathways in the astrocytic syncytium to focus local responses (Enkvist & McCarthy, 1992), our findings suggest that astrocytes better equipped for localized responses when presented with a stimulus during the active phase in mice. Conversely, during the rest period, characterized by more synchronous neuronal activity across broad brain areas (Vyazovskiy et al., 2009) higher GJC conductance might allow astrocytes to exert control over a larger area. In support of this idea, recent study showed that synchronized astrocytic Ca2+ activity advances the slow wave activity (SWA) of the brain, a key feature of non-REM sleep (Szabó et al., 2017). Blocking GJC was found to reduce SWA, further supporting this interpretation. However, conflicting findings have also been reported. For instance, Ingiosi et al. (Ingiosi et al., 2020) found that astrocytic synchrony was higher during wakefulness than sleep in the mouse frontal cortex. Whether these differing results in astrocyte synchrony during resting and active periods are attributable to differences in experimental context (e.g., brain regions, sleep-inducing condition) remains unclear. Indeed, astrocyte Ca2+ dynamics during wakefulness/sleep vary according to brain regions (Tsunematsu et al., 2021). While the extent of astrocyte synchrony might differ depending on brain region and/or stimulus, on our results suggest that the baseline state of astrocyte synchrony, which is affected by GJC conductance, varies with the day/night cycle.”

Reviewer #2 (Public Review):

Summary:

The article entitled "Circadian regulation of endoplasmic reticulum calcium response in mouse cultured astrocytes" submitted by Ryu and colleagues describes the circadian control of astrocytic intracellular calcium levels in vitro.

Strengths:

The authors used a variety of technical approaches that are appropriate

We appreciate the reviewer’s acknowledgement of the strengths of our manuscript.

Weaknesses:

Statistical analysis is poor and could lead to a misinterpretation of the data

Thank you for the comment. We have carefully reviewed our statistical analyses and applied appropriate methods where necessary. Please see below for the specific revisions and improvements made.

For Fig. 2D-E, we initially used a t-test. However, after adding more replicates and conducting a normality test, we found that the data did not follow a normal distribution. Therefore, we switched to the Mann-Whitney U test. In Fig. 5D-E, we originally used a repeated measures two-way ANOVA, but we have now changed it to a standard two-way ANOVA. For Fig. 7C and I, we also observed non-normal distribution in the normality test and consequently replaced the t-test with the Mann-Whitney U test. For other analyses not specifically mentioned, normality tests confirmed normal distribution, allowing us to use t-tests or ANOVA as appropriate for statistical analysis.

Several conceptual issues have been identified.

We have addressed the reviewer’s concerns. Please see our detailed point-by-point responses below.

Overinterpretation of the data should be avoided. This is a mechanistic paper done completely in vitro, all references to the in vivo situation are speculative and should be avoided.

We appreciate the reviewer’s insightful comment. Following the reviewer’s suggestion, we have removed the interpretations of GO pathways in the context of in vivo situation.

Reviewer #3 (Public Review):

Astrocyte biology is an active area of research and this study is timely and adds to a growing body of literature in the field. The RNA-seq, Herp expression, and Ca2+ release data across wild-type, Bmal1 knockout, and Herp knockdown cellular models are robust and lend considerable support to the study's conclusions, highlighting their importance. Despite these strengths, the manuscript presents a gap in elucidating the dynamics of HERP and the involvement of ITPR1/2 in modulating Ca2+ release patterns and their circadian variations, which remains insufficiently supported and characterized. While the Connexin data underscore the importance of rhythmic Ca2+ release triggered by ATP, the relationship here appears correlational and the role of HERP and ITPR in Cx function remains to be characterized. Moreover, enhancing the manuscript's clarity and readability could significantly benefit the presentation and comprehension of the findings.

We appreciate the reviewer’s acknowledgement of the strengths of our manuscript. Regarding the identified gaps, we have conducted several new experiments to clearly demonstrate the HERP-ITPR-Cx phosphorylation axis. Please see our detailed point-by-point responses below.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

- While HERP appears to be a clock-controlled gene and its protein levels appear to demonstrate rhythmicity as well, the data quality of the western blotting in Bmal1 knockout raises some concern about the accuracy of HERP protein quantification.

We understand the reviewer’s concern regarding the proximity of the HERP band to a nonspecific band in the Western blotting for the Bmal1 knockout. However, we took great care to ensure the accuracy of our HERP band quantification. We meticulously selected only the specific HERP band, excluding nonspecific band. Therefore, we are confident in the accuracy of our HERP protein measurements.

- If HERP is rhythmic and ITPRs are not, if their model is correct, might we expect HERP suppression to result in 'unmasking' an ITPR rhythm?

Our model suggests that both HERP and ITPRs are rhythmic, with HERP regulating the degradation of ITPR proteins and driving their rhythms. Consistent with this, we observed that day/night variations in ITPR2 levels (Fig. 4N and 4O). Therefore, we concluded that circadian variations in HERP are sufficient to drive ITPR2 rhythms. We have explained this in detail in the Result section (lines 236-241) and the Discussion section (lines 324-332).

- The authors make a rather abrupt switch to examining gap junctions and connexin 43 phosphorylation. While the data demonstrating that the phosphorylation of S368 may indeed be rhythmic - the authors do not connect these data to the rest of the manuscript by showing a connection to HERP-mediated calcium signaling, limiting the coherence of the narrative.

Thank you for the reviewer’s insightful comments. To address the reviewer's concern regarding the connection between Herp and the phosphorylation of CX43 at S368, we have conducted new experiments to test whether KD of Herp abolishes the rhythms of Cx43 phosphorylation at S368. We found that the phosphorylation of Cx43 at S368 is significantly enhanced at 30hrs post sync compared with 42hrs post sync in control siRNA-treated astrocytes consistent with our previous results (Fig. 6C & 6D). On the other hand, this circadian phase dependent difference in phosphorylation was abolished in Herp siRNA treated astrocytes. These results clearly indicate that circadian variations in Cx43 phosphorylation are driven by the HERP. These new results are now included in Fig. 6G and 6H and explained in the Results section (lines 276-281).

- Comment on data presentation: the authors repeatedly present histograms with attached lines between data points - from my understanding of the experiments, this is inappropriate unless these were repeated measures from the same cells. Otherwise, the lines connecting one data point to another between different conditions (e.g., Ctrl or Herp knockdown) are arbitrary and possibly misleading (i.e., Figure 3K, 3M, 4L, 6D).

Thank you for the reviewer’s comment. We have updated the figures by removing the lines connecting data points in the relevant figures (Fig.3K, M, Fig4.N and Fig.6D)).

Reviewer #2 (Recommendations For The Authors):

Most of the suggestions of this reviewer are related to the conceptual interpretation and presentation of the data and to the statistical analysis

In Figure 1 the authors analyzed the rhythmic transcriptome of cortical astrocytes synchronized with a serum shock in two different ways. The authors need to discuss what is the difference between the two methods used to detect rhythmic transcripts and make sense of them.

Following the reviewer’s suggestion, we have provided a more detailed explanation about MetaCycle and BioCycle, as well as the rationale for using both packages in our analysis as follows: “Various methods have been used to identify periodicity in time-series data, such as Lomb-Scargle (Glynn et al., 2006), JTK_CYCLE (Hughes et al., 2010) and ARSER (Yang & Su, 2010), each with distinct advantages and limitations. MetaCycle, integrates these three methods, facilitating the evaluation of periodicity in time-series data without requiring the selection of an optimal algorithm (Wu et al., 2016). Additionally, BioCycle has been developed using a deep neural network trained with extensive synthetic and biological time series datasets (Agostinelli et al., 2016). Because MetaCycle and Biocycle identify periodic signal based on different algorithms, we applied both packages to identify periodicity in our time-series transcriptome data. BioCycle and MetaCycle analyses detected 321 and 311 periodic transcripts, respectively (FDR corrected, q-value < 0.05) (Fig. 1B). Among these, 220 (53.4%) were detected by both methods, but many transcripts did not overlap. MetaCycle is known for its inability to detect asymmetric waveforms (Mei et al., 2020). In our analysis, genes with increasing waveforms like Adora1 and Mybph were identified as rhythmic only by BioCycle, while Plat and Il34 were identified as rhythmic only by MetaCycle (Fig. S1C). Despite these discrepancies, the clear circadian rhythmic expression profiles of these genes led us to conclude that using the union of the two lists compensates for the limitations of each algorithm.”

Please refer to lines 105-117 in the Results section.

The reasoning for comparing CT0 with the phase of the clock 8 hs after SS needs to be explained. Circadian time (CT) conceptually refers to the clock phase in the absence of entrainment cues in vivo, the direct transformation of "time after synchronization" in vitro to CT is misleading.

Thank you for the reviewer’s insightful comments. Initially, we believed that transforming TASS to CT, despite being in vitro data, might provide a more intuitive and physiologically relevant interpretation of our results. However, we agree that this approach might be misleading. Following the reviewer’s suggestion, we have revised our terminology by changing “CT” to “Time post sync (hr)”. Nonetheless, in Fig. 1F for circular peak phase map, we set 8hrs post sync to ZT0 based on a phase comparison result in Fig. 1D for physiologically relevant interpretation. We hope these revisions clarify our approach.

Moreover, also by definition a CT cannot be defined in terms of "dark" or "light". Figure 6M needs to be changed.

Following the reviewer’s suggestion, we removed the labels CT22 and CT34. Instead. we have labeled the respective periods as “30hr post sync” and “42hr post sync”.

In Figure 1D, the authors present a gene ontology analysis that is certainly interesting, however, it should not be overinterpreted when trying to explain processes that take place only in vivo (e.g. wound repair).

Thank you for the insightful comment. Following the reviewer’s feedback, we have removed the paragraph interpreting the cell migration process in relation to wound repair and have focused instead on Ca2+ ion homeostasis.

In Figure 2A the relative expression of clock genes and Herp is again misleading by a white/grey shading indicating subjective night and subjective day when the system under study is a cell culture.

We understand the reviewer’s concern that a cell culture system is not equivalent to light/dark entrainment condition. However, we apply time-synchronizing stimuli to recapitulate in vivo entrainment. In addition, by comparing our data with CircaDB, we defined 8hrs post sync as corresponding to ZT0, thus aligning it with the beginning of the day. We have retained the shading to facilitate easier interpretation of our data in relation to in vivo situations. However, in response to the reviewer’s concern, we have revised the shading from white/grey to light grey/dark grey. We hope this adjustment addresses the reviewer’s concern, but if the reviewer still believes it is inappropriate, please let us know, we will gladly update it.

In the Figure 2A legend, it is indicated that rhythmicity is assessed using MetaCycle with mean values obtained from n=2. The authors need to make clear whether this n=2 mean: 2 biological replicates or 2 technical replicates. This difference is relevant because it would make the analysis statistically valid or invalid, respectively.

Thank you for your feedback. n=2 refers to 2 biological replicates. Therefore, the analysis is statistically valid.

In Figures 2C and D the authors applied a T-test, a parametric statistical test for one-to-one comparison that requires normality distribution of the data to be tested first. To test normality, the authors need at least 4 biological replicates. The suggestion of this reviewer is that these experiments have to be repeated and proper statistics applied.

Thank you for your feedback. In response to the reviewer's suggestion, we conducted additional experiments to increase the number of biological replicates to 4. After verifying the normality of the data, we applied a t-test for Figure 2C and a Mann-Whitney test for Figure 2D and 2E. These tests confirmed significant statistical difference between groups.

Further evidence of Bmal1-dependent control of HERP circadian expression authors could check the presence of E-Box elements in the Herp promoter.

Thank you for the reviewer’s insightful comment. In the original version of our manuscript's Discussion section, we mentioned the absence of a canonical E-Box in the upstream of Herp gene. However, following the reviewer’s suggestion and considering the potential role of non-canonical E-Boxes, we conducted an additional analysis. This analysis identified several non-canonical E-Boxes within the 6 kb upstream region of the Herp gene (Table S2). Notably, we found one non-canonical E-Box, “CACGTT,” known to regulate circadian expression (Yoo et al., 2005) is close to the transcription start site (chr8:94386194-94386543). Moreover, this element is evolutionarily conserved across various mammals, including humans, rats, mice, dogs, and opossums (See Author response image 2). Therefore, we reasoned that these non-canonical E boxes might drive the CLOCK/BMAL1 dependent expression of Herp. We have updated the Discussion to reflect these findings in lines 315-319.

The calcium experiments shown in Figures 3A-I, could be more convincing if the authors showed that the different Ca2+ sensors are compartment-specific by showing co-localization with a subcellular marker. In the pictures shown it is not even possible to recognize the cell dimensions.

Following the reviewer’s suggestion, we performed co-staining experiments with organelle specific Ca2+ indicators and organelle markers. First, astrocytes were co-transfected with G-CEPIA1er, an ER specific Ca2+ indicator and ER targeted DsRed2 (with Calreticulin signal sequence). Live imaging analysis showed that the fluorescent intensities of G-CEPIA1er and DsRed2-ER-5 significantly overlapped in co-transfected cells. Secondly, astrocytes were transfected with Mito-R-GECO1 and Mitotracker, a cell permeable mitochondria dye, was applied. The fluorescent intensities of Mito-R-GECO1 and Mitotracker also significantly overlapped. These new data are included in Figure S4 and explained in the Result section (lines 194-195).

Data analysis in Figure 3 K and M is misleading. According to the explanations of the results, each of the experiments to assess ITRP1 or 2 is run independently. Then it is not clear why the relative levels obtained with control or Herp siRNA are plotted as pairs. Same comment as above for Figure 4L and Figure 6D.

Thank you for the reviewer’s insightful comments. Reviewer1 raised similar issues. Following the reviewers’ suggestions, we have removed the lines connecting the data points in Fig. 3K, 3M, 4L, and 6D.

In Figure 5E the authors need to explain why they consider that repeated measures 2-way ANOVA is the right statistical test to apply. According to the explained experimental design, cells transfected, synchronized, and then harvested independently at the indicated time after synchronization.

Thank you for the reviewer’s insightful comment. Upon reviewing the statistical methods as suggested, we have revised our approach. Instead of using repeated measures 2-way ANOVA, we have now applied a standard 2-way ANOVA, which is more appropriate given the experimental procedures were independent, as the reviewer pointed out.

The English language needs to be revised throughout the text.

We have thoroughly revised the English language throughout the text.

Reviewer #3 (Recommendations For The Authors):

(1) Figure 3. Clarify the physiological importance of 100 µM ATP. Would the Herp rhythm warrant Ca2+ release rhythms under basal conditions? In 3J-K, the relatively weak effect of Herp knockdown on ITPR1/2 levels, albeit statistically significant, may not be physiologically significant. This calls into question the claimed Herp-ITPR axis that underlies the Ca2+ release phenotype. Further, the correlation certainly exists but further characterization of Herp KD cells would be required to address the mechanism.

As previously reported, a broad range of ATP concentrations can induce Ca2+ activity in the astrocytes (Neary et al., 1988). Originally, we conducted an ATP dose-response analysis to observe ER Ca2+ release in our primary astrocyte culture. Our results show that ER Ca2+ release begins at 50 µM ATP and plateaus at 500 µM. Please refer to Author response image 3. We selected 100µM ATP for our experiments because it induces a medium level of ER Ca2+ response. Importantly, although measuring ATP concentrations at the synapse in vivo is challenging(Tan et al., 2017), estimates suggest synaptic ATP concentrations range from 5-500 µM (Pankratov et al., 2006). Thus, 100µM ATP is a physiologically relevant concentration that can affect nearby cells, including astrocytes, in the nervous system.

Cultured astrocytes were transfected with G-CEPIA1er ER and at 48hrs post transfection, cultured astrocytes were treated with various concentrations of ATP and Ca2+ imaging analysis was performed.

(A) ΔF/F0 values over time following ATP application. (B) Area above curve values. Values in graphs are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; one-way ANOVA).

Regarding the comment on Ca2+ release rhythms under basal conditions, we interpret this as referring Ca2+ release in the absence of a stimulus. We typically observe Ca2+ release only upon stimulation, such as ATP treatment. However, we acknowledge that the modest effects of HERP knockdown on ITPR1/2 levels could question the HERP-ITPR axis’s role in ER Ca2+ release.

To address this, we analyzed whether Herp KD induced increases in ER Ca2+ release were mediated through ITPRs by treating cells with Xestospongin C (XesC), an IP3R inhibitor. XesC treatment reduced ATP-induced ER Ca2+ release and eliminated the differences in ER Ca2+ release between control and Herp KD astrocytes (Fig. 3N – 3P). These results clearly indicate that HERP-ITPR axis plays critical role in controlling ER Ca2+ release. These new experiments have been included in Fig. 3 and explained in the result section (lines 217-221).

Furthermore, following the reviewer’s suggestion, we examined whether HERP rhythms underlie the rhythms of ER Ca2+ response by analyzing ER Ca2+ response in Herp KD astrocyte in two different times following synchronization. In control astrocytes, ATP-induced ER Ca2+ responses vary depending on time, whereas these time-dependent variations were abolished in Herp KD astrocytes. These new experiments have been included in Fig. 4K – 4M and explained in the Results section (lines 232-235).

Collectively, these results indicate that HERP rhythms lead to time-dependent differences in ER Ca2+ response through ITPRs.

(2) Figure 4K-L. As data suggested the involvement of ITPR1 and ITPR2 (circadian effect), a reasonable next step is to determine their involvement, but the study did not pursue the hypothesis.

Thank you for your insightful comment. Our results indeed suggest that rhythms in ITPR2 levels may drive the time-dependent variations in ATP-induced ER Ca2+ release following synchronization. The newly conducted experiments demonstrated that treatment with the ITPR inhibitor XesC suppressed ATP-induced ER Ca2+ release at both control and Herp siRNA treatment conditions (Fig. 3). Based on these findings, we now further confirm that rhythms of ITPR levels, specifically ITPR2 underlie the circadian variations in ER Ca2+ release. While examining the effect of ITPR2 siRNA would directly prove the involvement of ITPR2, we have decided to pursue this experiment in the future studies.

(3) Figure 5A-C. Data from WT cells should be included side by side with Bmal1-/- cells for comparison which is expected to be consistent with the HERP levels as in 5D-E. Again, the role of ITPR2 is suggested but not demonstrated.

Following the reviewer's suggestion, we conducted additional experiments including both WT and Bmal1-/- cultured astrocytes side-by-side. The results were consistent with our previous findings: WT astrocytes showed rhythms of ER Ca2+ release while Bmal1-/- astrocytes did not. We have updated the Figure 5A to 5C and the corresponding Results section in lines 242-245 accordingly.

Regarding second comment, as mentioned in our previous response, we plan to examine the role of ITPR2 in further studies.

(4) Figure 6. The Connexin data seems an addon and is correlative with the Ca2+ release. The role of Herp and Itpr in Connexin function is not addressed. Figure 6E-F was not called out in the results section. Suggest providing additional data to support the role of the HERP-ITPR axis in regulating Ca2+ release and Connexin activity.

We agree that additional data are needed to support the role of HERP in regulating CX43 phosphorylation. Therefore, we have conducted further experiments to determine whether rhythms of Cx43 phosphorylation are regulated by HERP. In the control astrocytes, ATP treatment induced time-dependent variations in Cx43 phosphorylation. However, these rhythms were abolished in Herp KD astrocytes. These results indicate that rhythms in HERP levels contribute to the time-dependent variations in Cx43 phosphorylation. These new experiments have included in Fig. 6G and 6H and explained in the results section (lines 276-281).

Regarding second comment, we have corrected our oversight by properly referencing figures 6E-F in the results section. Please refer to lines 357-359 for clarification.

(5) Discussion. This section should focus on noteworthy points to discuss, not repeating the results.

Based on the reviewer's valuable suggestions, we have revised the Discussion section to minimize repetition of the results. Thank you for your guidance.

(6) The manuscript exhibits numerous grammatical and textual inaccuracies that necessitate careful revision by the authors. My observations here are confined to the title and the abstract alone. I recommend altering the title from "mouse cultured astrocytes" to "cultured mouse astrocytes" for clarity and grammatical correctness. The abstract, meanwhile, needs enhancements both in terms of its content and language. It should incorporate the results of the partitioning among the ER, cytoplasm, and mitochondria, and provide clear definitions for some of the critical terms used. It's worth noting that the abstract's second sentence contains a grammatical error.

Thank you for the reviewer’s valuable feedback. We have carefully revised the title, abstract, and main text to address the grammatical and textual issues. The title has been changed to “cultured mouse astrocytes”. Additionally, the abstract now includes results related to cytoplasmic Ca2+ dynamics and has been revised in several places. We appreciate your insights and have worked to enhance the content and language accordingly.

Reference

Agostinelli, F., Ceglia, N., Shahbaba, B., Sassone-Corsi, P., & Baldi, P. (2016). What time is it? Deep learning approaches for circadian rhythms. Bioinformatics, 32(12), i8-i17. https://doi.org/10.1093/bioinformatics/btw243

Cahoy, J. D., Emery, B., Kaushal, A., Foo, L. C., Zamanian, J. L., Christopherson, K. S., Xing, Y., Lubischer, J. L., Krieg, P. A., Krupenko, S. A., Thompson, W. J., & Barres, B. A. (2008). A transcriptome database for astrocytes, neurons, and oligodendrocytes: a new resource for understanding brain development and function. J Neurosci, 28(1), 264-278. https://doi.org/10.1523/JNEUROSCI.4178-07.2008

Carreras-Sureda, A., Pihán, P., & Hetz, C. (2018). Calcium signaling at the endoplasmic reticulum: fine-tuning stress responses. Cell Calcium, 70, 24-31. https://doi.org/10.1016/j.ceca.2017.08.004

Enkvist, M. O., & McCarthy, K. D. (1992). Activation of protein kinase C blocks astroglial gap junction communication and inhibits the spread of calcium waves. J Neurochem, 59(2), 519-526. https://doi.org/10.1111/j.1471-4159.1992.tb09401.x

Fujii, Y., Maekawa, S., & Morita, M. (2017). Astrocyte calcium waves propagate proximally by gap junction and distally by extracellular diffusion of ATP released from volume-regulated anion channels. Scientific Reports, 7(1), 13115. https://doi.org/10.1038/s41598-017-13243-0

Giorgi, C., Marchi, S., & Pinton, P. (2018). The machineries, regulation and cellular functions of mitochondrial calcium. Nature Reviews Molecular Cell Biology, 19(11), 713-730. https://doi.org/10.1038/s41580-018-0052-8

Glynn, E. F., Chen, J., & Mushegian, A. R. (2006). Detecting periodic patterns in unevenly spaced gene expression time series using Lomb-Scargle periodograms. Bioinformatics, 22(3), 310-316. https://doi.org/10.1093/bioinformatics/bti789

Hughes, M. E., Hogenesch, J. B., & Kornacker, K. (2010). JTK_CYCLE: an efficient nonparametric algorithm for detecting rhythmic components in genome-scale data sets. J Biol Rhythms, 25(5), 372-380. https://doi.org/10.1177/0748730410379711

Ingiosi, A. M., Hayworth, C. R., Harvey, D. O., Singletary, K. G., Rempe, M. J., Wisor, J. P., & Frank, M. G. (2020). A Role for Astroglial Calcium in Mammalian Sleep and Sleep Regulation. Curr Biol, 30(22), 4373-4383.e4377. https://doi.org/10.1016/j.cub.2020.08.052

Mei, W., Jiang, Z., Chen, Y., Chen, L., Sancar, A., & Jiang, Y. (2020). Genome-wide circadian rhythm detection methods: systematic evaluations and practical guidelines. Briefings in Bioinformatics, 22(3). https://doi.org/10.1093/bib/bbaa135

Neary, J. T., van Breemen, C., Forster, E., Norenberg, L. O., & Norenberg, M. D. (1988). ATP stimulates calcium influx in primary astrocyte cultures. Biochem Biophys Res Commun, 157(3), 1410-1416. https://doi.org/10.1016/s0006-291x(88)81032-5

Pankratov, Y., Lalo, U., Verkhratsky, A., & North, R. A. (2006). Vesicular release of ATP at central synapses. Pflugers Arch, 452(5), 589-597. https://doi.org/10.1007/s00424-006-0061-x

Paredes, F., Parra, V., Torrealba, N., Navarro-Marquez, M., Gatica, D., Bravo-Sagua, R., Troncoso, R., Pennanen, C., Quiroga, C., Chiong, M., Caesar, C., Taylor, W. R., Molgó, J., San Martin, A., Jaimovich, E., & Lavandero, S. (2016). HERPUD1 protects against oxidative stress-induced apoptosis through downregulation of the inositol 1,4,5-trisphosphate receptor. Free Radic Biol Med, 90, 206-218. https://doi.org/10.1016/j.freeradbiomed.2015.11.024

Szabó, Z., Héja, L., Szalay, G., Kékesi, O., Füredi, A., Szebényi, K., Dobolyi, Á., Orbán, T. I., Kolacsek, O., Tompa, T., Miskolczy, Z., Biczók, L., Rózsa, B., Sarkadi, B., & Kardos, J. (2017). Extensive astrocyte synchronization advances neuronal coupling in slow wave activity in vivo. Scientific Reports, 7(1), 6018. https://doi.org/10.1038/s41598-017-06073-7

Tan, Z., Liu, Y., Xi, W., Lou, H. F., Zhu, L., Guo, Z., Mei, L., & Duan, S. (2017). Glia-derived ATP inversely regulates excitability of pyramidal and CCK-positive neurons. Nat Commun, 8, 13772. https://doi.org/10.1038/ncomms13772

Torrealba, N., Navarro-Marquez, M., Garrido, V., Pedrozo, Z., Romero, D., Eura, Y., Villalobos, E., Roa, J. C., Chiong, M., Kokame, K., & Lavandero, S. (2017). Herpud1 negatively regulates pathological cardiac hypertrophy by inducing IP3 receptor degradation. Sci Rep, 7(1), 13402. https://doi.org/10.1038/s41598-017-13797-z

Tsunematsu, T., Sakata, S., Sanagi, T., Tanaka, K. F., & Matsui, K. (2021). Region-specific and state-dependent astrocyte Ca2+ dynamics during the sleep-wake cycle in mice. The Journal of Neuroscience, JN-RM-2912-2920. https://doi.org/10.1523/jneurosci.2912-20.2021

Verkhratsky, A., & Nedergaard, M. (2018). Physiology of Astroglia. Physiol Rev, 98(1), 239-389. https://doi.org/10.1152/physrev.00042.2016

Vyazovskiy, V. V., Olcese, U., Lazimy, Y. M., Faraguna, U., Esser, S. K., Williams, J. C., Cirelli, C., & Tononi, G. (2009). Cortical firing and sleep homeostasis. Neuron, 63(6), 865-878. https://doi.org/10.1016/j.neuron.2009.08.024

Wu, G., Anafi, R. C., Hughes, M. E., Kornacker, K., & Hogenesch, J. B. (2016). MetaCycle: an integrated R package to evaluate periodicity in large scale data. Bioinformatics, 32(21), 3351-3353. https://doi.org/10.1093/bioinformatics/btw405

Yang, R., & Su, Z. (2010). Analyzing circadian expression data by harmonic regression based on autoregressive spectral estimation. Bioinformatics, 26(12), i168-174. https://doi.org/10.1093/bioinformatics/btq189

Yoo, S. H., Ko, C. H., Lowrey, P. L., Buhr, E. D., Song, E. J., Chang, S., Yoo, O. J., Yamazaki, S., Lee, C., & Takahashi, J. S. (2005). A noncanonical E-box enhancer drives mouse Period2 circadian oscillations in vivo. Proc Natl Acad Sci U S A, 102(7), 2608-2613. https://doi.org/10.1073/pnas.0409763102

Version 2

Abstract

Abstract

The circadian clock, an internal time-keeping system orchestrates 24-hour rhythms in physiology and behavior by regulating rhythmic transcription in cells. Astrocytes, the most abundant glial cells, play crucial roles in central nervous system (CNS) functions, but the impact of the circadian clock on astrocyte functions remains largely unexplored. In this study, we identified 412 circadian rhythmic transcripts in cultured mouse cortical astrocytes through RNA sequencing. Gene Ontology analysis indicated that genes involved in Ca2+ homeostasis are under circadian control. Notably, Herpud1 (Herp) exhibited robust circadian rhythmicity at both mRNA and protein levels, a rhythm disrupted in astrocytes lacking the circadian transcription factor, BMAL1. HERP regulated endoplasmic reticulum (ER) Ca2+ release by modulating the degradation of inositol 1,4,5-trisphosphate receptors (ITPRs). ATP-stimulated ER Ca2+ release varied with the circadian phase, being more pronounced at subjective night phase, likely due to the rhythmic expression of ITPR2. Correspondingly, ATP-stimulated cytosolic Ca2+ increases were heightened at the subjective night phase. This rhythmic ER Ca2+ response led to circadian phase-dependent variations in the phosphorylation of Connexin 43 (Ser368) and gap junctional communication. Given the role of gap junction channel (GJC) in propagating Ca2+ signals, we suggest that this circadian regulation of ER Ca2+ responses could affect astrocytic modulation of synaptic activity according to the time of day. Overall, our study enhances the understanding of how the circadian clock influences astrocyte function in the CNS, shedding light on their potential role in daily variations of brain activity and health.

Introduction

The circadian clock orchestrates a range of behaviors, physiological processes, and cellular and biochemical activities in organisms with approximately 24-hour rhythms, facilitating adaptation to environmental changes (Bell-Pedersen et al., 2005). In mammals, the master clock localized in the suprachiasmatic nucleus (SCN) of the anterior hypothalamus synchronizes peripheral tissue clocks throughout the body to the light/dark cycle (Mohawk et al., 2012). The circadian clock’s molecular mechanism is driven by a transcriptional translational feedback loop (TTFL). CLOCK and BMAL1 form a heterodimer that triggers expression of the repressor genes, Period (Per) and Cryptochrome (Cry), along with other clock-controlled output genes. PER and CRY proteins then translocate to the nucleus, where they interact with CLOCK/BMAL1, halting further transcriptional activation. A secondary feedback loop involving the interplay between transcriptional activation by ROR and repression by REV-ERB finely regulates the rhythmic transcription of Bmal1 (Takahashi, 2017). Recent studies have revealed that the circadian clock controls the rhythmic transcription of numerous genes, displaying notable tissue specificity. This characteristic underscores the time-of-day-specific functional specialization of diverse tissues (Mure et al., 2018; Zhang et al., 2014). Consequently, circadian clock-controlled genes significantly influence various physiological processes including development, metabolism, aging, and neurodegeneration (Lananna & Musiek, 2020; Reinke & Asher, 2019; Schultz & Kay, 2003).

Astrocytes, the most abundant glial cells in the CNS, play vital roles in maintaining homeostasis of extracellular fluids, ions, and neurotransmitters; proving energy substrates to neurons; regulating local blood flow; modulating neuronal activity through gliotransmission; and assisting in interstitial fluid drainage (Sofroniew & Vinters, 2010). Though not electrically excitable, astrocytes exhibit complex intracellular Ca2+ signaling, both spatially and temporally, within individual astrocytes and across astrocytic networks. Astrocytes express various types of Gq-coupled receptors activated by neurotransmitters such as ATP, epinephrine, glutamate, and dopamine etc. (Verkhratsky & Nedergaard, 2018). Activation of these Gq-coupled receptors initiates the inositol 1,4,5-trisphosphate receptor (IP3R) signaling cascade, causing a release of Ca2+ from the ER that substantially increases intracellular Ca2+ levels (Guerra-Gomes et al., 2017; Verkhratsky & Nedergaard, 2018). Increases in intracellular Ca2+ in astrocytes are involved in regulating numerous afore-mentioned processes (Giaume et al., 2021; Verkhratsky & Nedergaard, 2018). Recent studies have shown that astrocyte Ca2+ signaling can regulate circadian rhythms and sleep patterns (Bojarskaite et al., 2020; Brancaccio et al., 2019).

Accumulating evidence underscores the importance of the circadian clock in astrocytes function. Cultured murine cortical astrocytes expressing a PER-luciferase reporter (PER::LUC) exhibit circadian oscillations and can be entrained by temperature cycle (Prolo et al., 2005) and vasoactive intestinal polypeptide (VIP), which synchronizes SCN (Marpegan et al., 2009). These findings provide strong evidence that a molecular clock operates in astrocytes. In addition, disruption of clock genes specific to astrocytes has been linked to a range of deleterious effects, including excessive astrocyte activation, disruptions in daily activity patterns, impaired learning, metabolic imbalances, and a shortened lifespan (Barca-Mayo et al., 2020; Barca-Mayo et al., 2017; Griffin et al., 2019; Lananna et al., 2018). Remarkably, these phenotypic changes closely resemble those observed in astrocytes during the progression of neurodegenerative diseases (Brandebura et al., 2023; Phatnani & Maniatis, 2015). Moreover, the rhythmic metabolic activity of astrocytes, including intracellular Ca2+ fluctuations and neurotransmitter release, is critical for circadian timing in the SCN (Brancaccio et al., 2017). Despite the emerging significance of circadian rhythms in astrocyte function in health and disease, our understanding of the link between the circadian clock and specific astrocytic functions is still limited.

In this study, we conducted a transcriptome analysis to investigate how the circadian clock regulates astrocyte function. Our research uncovered that Herp (homocysteine-inducible ER protein with ubiquitin like domain 1; Herp) exhibited a BMAL1-dependent rhythmic expression, peaking at subjective midday phase and reaching its nadir at subjective midnight phase. HERP was crucial for ER Ca2+ regulation by modulating the degradation of ITPRs. Importantly, the ATP-stimulated ER Ca2+ response varied according to time post synchronization, a dependence that was absent in cultured astrocytes from Bmal1-/-mice. Furthermore, the pronounced release of Ca2+ during the subjective night phase led to robust phosphorylation of the Ser368 residue of CX43, diminishing cell-to-cell communication between astrocytes. Overall, our study demonstrates that the circadian clock orchestrates a variety of astrocytic processes including Ca2+ homeostasis by controlling the rhythmic transcription of genes, highlighting its pivotal role in CNS function.

Results

Defining circadian rhythmic transcripts in cultured mouse cortical astrocytes

To understand how the circadian clock regulates astrocyte physiology, we conducted a circadian transcriptome analysis of cultured mouse cortical astrocytes. Cultured astrocytes were synchronized using serum shock (SS) and harvested every 4hrs for 2 days starting at 12hrs post-SS. The time series astrocyte transcriptome was determined by RNA-seq (Figure 1A). Using a Gaussian mixture model, we defined expressed transcripts, choosing 0.577 transcripts per million (TPM) as the cut-off, corresponding to the 1% threshold in the distribution curve of highly expressed genes (Figure 1-figure supplement 1A). A total of 17,671 transcripts with the highest TPM values of the 12 time points exceeding 0.577, were considered expressed. To validate astrocyte enrichment, we compared the expression of marker genes for nervous cells. Our transcriptome data were replete with astrocyte markers but lacked markers for microglia, oligodendrocytes, neurons, and endothelial cells (Figure 1-figure supplement 1B).

Circadian rhythmic transcripts in mouse cultured cortical astrocytes.

(A) Experimental scheme for synchronizing circadian rhythms in mouse cultured cortical astrocytes followed by RNA-sequencing over 2 days. (B) Venn diagram displaying the number of circadian rhythmic transcripts identified by two algorithms (q < 0.05 in MetaCycle or BioCycle). (C) Number of transcripts that overlapping with 12 tissues in mouse circadian transcriptome ss from CircaDB (http://circadb.hogeneschlab.org). (D) Comparison of mean phase (ZT) from CircaDB and peak phase (Time post sync) from cultured astrocytes for 14 transcripts that are rhythmic in 10 or more tissues. (E) Scatter plot showing phase in CircaDB and peak phase of cultured astrocytes for transcripts that are rhythmic in one tissue. (F) Radial histogram of the distribution of phases of rhythmic genes in the astrocyte transcriptome. (G) Top 10 enriched GO Biological Process (BP) terms for significant circadian rhythmic genes in astrocytes (p < 0.01) identified by Metascape (https://metascape.org).

Various methods have been used to identify periodicity in time-series data, such as Lomb-Scargle (Glynn et al., 2006), JTK_CYCLE (Hughes et al., 2010) and ARSER (Yang & Su, 2010), each with distinct advantages and limitations. MetaCycle, integrates these three methods, facilitating the evaluation of periodicity in time-series data without requiring the selection of an optimal algorithm (Wu et al., 2016). Additionally, BioCycle has been developed using a deep neural network trained with extensive synthetic and biological time series datasets (Agostinelli et al., 2016). Because MetaCycle and Biocycle identify periodic signal based on different algorithms, we applied both packages to identify periodicity in our time-series transcriptome data. BioCycle and MetaCycle analyses detected 321 and 311 periodic transcripts, respectively (FDR corrected, q-value < 0.05) (Figure 1B). Among these, 220 (53.4%) were detected by both methods, but many transcripts did not overlap. MetaCycle is known for its inability to detect asymmetric waveforms (Mei et al., 2020). In our analysis, genes with increasing waveforms like Adora1 and Mybph were identified as rhythmic only by BioCycle, while Plat and Il34 were identified as rhythmic only by MetaCycle (Figure 1-figure supplement 1C). Despite these discrepancies, the clear circadian rhythmic expression profiles of these genes led us to conclude that using the union of the two lists compensates for the limitations of each algorithm. Consequently, we identified a total of 412 circadian rhythmic transcripts (2.3% of all transcripts) (Supplementary file 1). The circadian oscillations of these transcripts are illustrated in the heatmap in Figure 1-figure supplement 1D.

Next, we compared rhythmic transcripts of astrocytes with those of 12 mouse tissues from CircaDB (Zhang et al., 2014), a database of circadian rhythmic genes (Figure 1C). A majority of the rhythmic transcripts in astrocytes (265 of 412, 65.3%) were not rhythmic or were rhythmic in one tissue. 14 transcripts were rhythmic in more than 10 tissues; most of these were core clock genes except Tspan4, Tsc22d3, and Wee1. This result is consistent with previous reports of tissue-specific circadian rhythmic expression (Mure et al., 2018; Panda et al., 2002). A phase comparison of these 14 common oscillating transcripts between CircaDB and our analysis showed a consistent time difference, with 8hrs after SS corresponding to ZT0 (Figure 1D). An additional phase comparison of 82 transcripts that overlapped in only one tissue indicated a very robust phase correlation between astrocytes and the corresponding tissue (r = 0.69, p < 0.001) (Figure 1E). Collectively, our analysis indicates that the identified oscillating transcripts were not false positive but were clearly associated with the circadian clock. Hierarchical clustering analysis of oscillating transcripts revealed two clusters: cluster 1 with 185 transcripts and cluster 2 with 227 transcripts (Figure 1-figure supplement 2). A circular map of peak phases across all rhythmic genes revealed two major peaks, one in the early night (ZT12 to ZT16) and one in the late night or dawn (ZT21 to ZT1) (Figure 1F). Here for more intuitive interpretation, we set 8hrs post synchronization (sync) to ZT0 based on a previous phase comparison. Circadian transcriptome studies conducted in multiple animal tissues have indicated that the phases of each tissue are not randomly distributed but are predominantly clustered in one or two narrow temporal windows (Panda et al., 2002). Consistent with these previous reports, mouse cultured astrocytes exhibited a high degree of temporal organization. Intriguingly, there was a relatively quiescent zone during subjective daytime, a finding different from that observed for diurnal animals, which showed a quiescent zone during subjective nighttime (Mure et al., 2018).

To understand the biological processes and pathways controlled by the circadian clock, we performed a Gene Ontology (GO) analysis on each cluster using Metascape (https://metascape.org) (Figure 1G). Given that the goal of our research was to investigate processes specifically controlled by the circadian clock, we excluded 17 core clock transcripts (Arntl, Clock, Nfil3, Npas2, Bhlhe40, Bhlhe41, Cry1, Cry2, Dbp, Elf, Nr1d1, Nr1d2, Per1, Per2, Per3, Rorc, and Tef) from our GO enrichment analysis. Among the rhythmic processes, we focused on “calcium ion homeostasis (GO:0055074)” due to the importance of intracellular Ca2+ signaling in numerous astrocyte functions (Agulhon et al., 2008; Bojarskaite et al., 2020; Brancaccio et al., 2019; Giaume et al., 2021; Guerra-Gomes et al., 2017; Verkhratsky & Nedergaard, 2018). In this category, Ca2+ homeostasis-related genes included Herp, Slc4a11, Sord and Kcnh1 etc. with Herp showing the most robust oscillation with large amplitude (Table S1).

Circadian rhythmic expression of Herp in cultured mouse cortical astrocytes

We first assessed the expression patterns of core clock genes and Herp in our cultured astrocytes. Bmal1, was rhythmically expressed with a peak at 30hr post sync while Per2 and Rev-Erbα showed rhythmic expression with peaks at 20hr and 16hr, respectively (Figure 2A). Herp displayed a robust rhythmic expression pattern, peaking at 20hr, similar to Per2 (Figure 2A). To verify the circadian expression of Herp, we performed quantitative real-time RT-PCR, which confirmed that Bmal1, Rev-Erbα and Herp exhibited circadian rhythms consistent with the RNAseq results (Figure 2B).

Herp is rhythmically expressed in mouse cultured astrocytes and its expression is controlled by BMAL1.

(A, B) Astrocyte cultures from rom WT (black) and Bmal1-/-(orange) mice were synchronized and harvested at the indicated time. Expression was analyzed for rhythmicity using MetaCycle, and expression levels of given genes were quantified using RNA-seq (A) and real-time qRT-PCR (B) data. Values are mean ± SEM (n = 2; p-values are indicated by insets in graphs). Light gray and dark gray backgrounds represent subjective day and subjective night, respectively, based on the phase analysis from Figure 1D. (C-E) Cultured astrocytes were transfected with the indicated siRNA (20nM) and processed for real-time qRT-PCR (C) and Western blot (D, E) analyses at 48hrs post transfection. (C) Values are mean ± SEM (n = 4; *p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; t-test). (D) Representative Western blot images of four independent experiments are shown. GAPDH served as loading control. (E) Densitometric quantification of HERP levels, normalized to GAPDH levels. Values are mean ± SEM (n = 4; ***p < 0.0005; Mann-Whiteny U test). (F-I) Astrocyte cultures from rom WT (F, G) and Bmal1-/- (H, I) mice were synchronized and harvested at the indicated time for Western blot analysis. Total ERK (tERK) served as a loading control. HERP/tERK values at different times were normalized to those at 12hr post sync (set to 1). Meta2d p values are indicated by insets in graphs. Light gray and dark gray backgrounds represent subjective day and subjective night, respectively. (F) Representative Western blot images from five independent experiments are shown. (H) Representative Western blot images from two independent experiments are shown

Next, we examined whether HERP protein levels also showed rhythmic changes, as many genes exhibit rhythmic transcription without corresponding protein level changes (Reddy et al., 2006). We validated the specificity of HERP antiserum using Western blot analysis of astrocytes treated with Herp siRNA (Herp-KD), which significantly reduced Herp mRNA levels (Figure 2C) and the intensities of specific bands in Western blots (Figure 2D and 2E). SS-synchronized cultured astrocytes were sampled every 6hrs for 2 days and processed for Western blot analysis of BMAL1 and HERP. BMAL1 phosphorylation showed circadian variation, with hyper-phosphorylation peaking at midday and hypo-phosphorylation peaking early in the day or late at night (Yoshitane et al., 2009). In our SS-synchronized cultured astrocytes, BMAL1 was hyperphosphorylated at 18hr, 24hr, 42hr, and 48hr (corresponding to midday) and hypo-phosphorylated at 12hr, 36hr (corresponding to early in the day), and 30hr, 54hr (corresponding to late at night) throughout two daily cycles (Figure 2F) indicating strong circadian rhythms in cultured astrocytes. Importantly, HERP protein levels oscillated peaking at 18hr and 42hr and reaching a minimum at 12hr, 30hr, and 54hr, mirroring its mRNA pattern (Figure 2F and 2G).

Herp also exhibited rhythmic expression in astrocyte cultures synchronized by either dexamethasone (Balsalobre et al., 2000) or forskolin (Yagita & Okamura, 2000) treatments which are other commonly used for circadian clock resetting (Figure 2-figure supplement 1). Finally, to investigate whether the rhythmic expression of Herp is regulated by a circadian clock, we examined Herp mRNA and protein expression patterns in primary astrocyte cultured from Bmal1-/- mice. The rhythmic expression patterns of Per2, Rev- Erbα, and Herp were abolished in Bmal1-/- astrocyte cultures with their expression maintained at trough levels (Figure 2A). HERP protein levels remained constant throughout the daily cycles (Figure 2H and 2I). Collectively, these observations indicate that the expression of Herp exhibits a robust circadian rhythm that is controlled by BMAL1 in cultured mouse astrocytes. These results suggest that HERP-regulated cellular processes vary in a time-of-day specific manner.

Herp knockdown alters ATP-induced ER Ca2+ release

Herp was first identified as a gene with altered expression in response to homocysteine treatment in human umbilical vein endothelial cells (HUVECs) (Kokame et al., 1996). It is strongly induced not only by homocysteine but also by ER stress-causing agents such as tunicamycin or thapsigargin. HERP, characterized by an N-terminal ubiquitin-like domain and present in the ER membrane facing cytoplasmic side, is involved in the unfolded protein response (UPR) (Kokame et al., 2000) and is part of the ER- associated degradation (ERAD) complex, participating in the ubiquitination and relocation of ERAD substrates (Leitman et al., 2014). Notably, HERP modulates the ER Ca2+ response through IP3R degradation (Paredes et al., 2016; Torrealba et al., 2017).

Thus, we sought to investigate whether HERP-controlled processes are subject to circadian regulation by examining the HERP-regulated ER Ca2+ response in the astrocytes. We employed the organelle-specific fluorescent Ca2+ indicators: G-CEPIA1er for the ER (Suzuki et al., 2014), R-GECO1 for the cytosol (Wu et al., 2013), and mito-R-GECO1 for mitochondria (Zhao et al., 2011). We first confirmed that these Ca2+ indicators colocalized with organelle-specific markers (Figure 3-figure supplement 1). ATP serves not only as a fundamental energy source but also as an active intercellular messenger. In astrocytes, ATP application induces an increase in intracellular Ca2+ (Neary et al., 1988) by binding to the Gq/G11 coupled P2Y receptor, activating PLC-β and inducing the hydrolysis of PIP2 to diacylglycerol (DAG) and IP3, which mobilizes Ca2+ through IP3Rs on the ER membrane (Baryshnikov et al., 2003; James & Butt, 2001).

We treated control and Herp-KD astrocytes with 100μM ATP and monitored subcellular Ca2+ changes. ATP treatment rapidly decreased ER Ca2+ in control astrocytes and this response was more pronounced in Herp-KD astrocytes (Figure 3A-3C). As the main Ca2+ store in the cell, ER releases Ca2+ which is then rapidly transmitted to other organelles such as mitochondria and cytosol (Carreras-Sureda et al., 2018; Giorgi et al., 2018). Although ATP treatment did not significantly alter cytosolic Ca2+ signal in control astrocytes, it greatly increased cytosolic Ca2+ signals in Herp-KD astrocytes (Figure 3D-3F). Consistently, mitochondrial Ca2+ significantly increased in Herp-KD astrocytes compared with control (Figure 3G-3I).

Herp knockdown altered ATP-induced ER Ca2+ response.

(A-I) Cultured astrocytes were co-transfected with 20 μM non-targeting (CTRL) siRNA or Herp siRNA together with G-CEPIA1er (A-C), R-GECO1 (D-F) or mito-R-GECO1 (G-I). At 48hrs post transfection, cultured astrocytes were treated with 100 µM ATP and Ca2+ imaging analysis was performed. Images were acquired every 3 seconds. (A, D, G) Representative time-lapse images of each Ca2+ indicator. (B, E, H) ΔF/F0 values over time following ATP application. (C, F, I) Area above or area under the curve values, calculated from panels B, E, and H. (A-C) CTRL siRNA, n = 19; Herp siRNA, n = 22. (D-F) CTRL siRNA, n = 20; Herp siRNA, n = 25. (G-I) CTRL siRNA, n = 16; Herp siRNA, n = 16. (J-M) Cultured astrocytes were transfected with the indicated siRNA (20nM) and processed for Western blot analysis 48hrs post transfection. Vinculin and GAPDH served as loading control for ITPRs and HERP, respectively. (J) Representative Western blot images from twelve independent experiments are shown. NS, non-specific band (K) Densitometric quantification of Western blot data showing relative levels of ITPR1 in Herp siRNA-transfected astrocytes compared to CTRL siRNA transfected astrocytes. (L) Representative Western blot images from five independent experiments. (M) Densitometric quantification of Western blot data showing relative levels of ITPR2 in Herp siRNA-transfected astrocytes compared to Control astrocytes. Values are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; t-test). (N-P) Cultured astrocytes were treated with 10μM Xestospongin C (XesC), an IP3R inhibitor, for 30 minutes before live imaging. Cells were then treated with 100μM ATP, and images were captured every 3 seconds. (N) Representative time-lapse images of ER Ca2+ indicator. (O) ΔF/F0 values over time following ATP application. (P) Area above the curve values were calculated from panel B. CTRL siRNA + Mock, n = 9; Herp siRNA + Mock, n = 9 ; CTRL siRNA + XesC, n = 8; Herp siRNA+ XesC, n = 14. Values are means ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; one-way ANOVA).

Next, we examined the mechanism by which HERP controls ER Ca2+.There are three subtypes of IP3Rs encoded by the genes Itpr1, Itpr2, and Itpr3. Transcriptome analysis of our cultured astrocytes indicated Itpr1 as the most prevalent, followed by Itpr2, while Itpr3 expression was minimal (Figure 3-figure supplement 2). Notably, expression of these Itpr subtypes was not rhythmic. We then assessed the ITPR1 and ITPR2 protein levels in control and Herp-KD astrocytes and found that the levels of both ITPR1 and ITPR2 were slightly but statistically significantly increased in Herp-KD astrocytes compared to controls (Figure 3J-3M). These results, consistent with previous reports (Paredes et al., 2016; Torrealba et al., 2017) indicated that HERP negatively regulates ITPRs in astrocyte.

To confirm that the HERP-ITPR axis underlies ER Ca2+ response, we analyzed the ER Ca2+ response following treatment with Xestospongin C (XesC), an IP3R inhibitor. XesC treatment reduced the ATP- induced ER Ca2+ release and abolished the enhanced effect observed in Herp-KD compared to control (Figure 3N-3P). Collectively, these results clearly indicate that HERP controls ER Ca2+ release by negatively regulating IP3Rs.

ATP-induced ER Ca2+ release varies according to time post synchronization

Given that HERP levels exhibited circadian variation and HERP regulated ER Ca2+ release, we investigated whether ATP-induced ER Ca2+ responses differ depending on circadian phase. After SS, we measured ER Ca2+ following ATP treatment at peak (42hr post sync) and trough (30hr post sync) phases of HERP (Figure 4A). With lower HERP levels potentially resulting in higher IP3R levels at 30hr, we anticipated greater ER Ca2+ release at 30hr than at 42hr. Consistent with this, ER Ca2+ decreased more at 30hr than at 42hr upon ATP treatment (Figure 4B-4D). Also, in keeping with previous result (Figure 3A-3I), cytosolic Ca2+ increased more at 30hr than at 42hr following ATP treatment (Figure 4E-4G). Although mitochondrial Ca2+ responses exhibited a similar trend, the differences were not statistically significant (Figure 4H-4J).

ATP-induced ER Ca2+ release varies according to time post sync.

(A) Schematic diagram of the experimental scheme from transfection to live-cell Ca2+ imaging at different times. (B-M) Cultured astrocytes were transfected with G-CEPIA1er (B-D, K-M), R-GECO1 (E-G), or mito-R-GECO1 (H-J) compartment-specific Ca2+ indicators (denoted at left) and then their circadian rhythm was synchronized by SS. (K-M) The indicated siRNA was co-transfected with the ER Ca2+ indicator. After transfection, cells were allowed 48hrs for the siRNA to take effect and stabilize before synchronization by serum shock. At the indicated times, astrocytes were treated with 100 µM ATP and Ca2+ imaging was performed. (B, E, H, K) Representative time-lapse images of each Ca2+ indicator. (C, F, I, L) ΔF/F0 values over time following ATP application. (D, G, J, M) Area above or area under the curve values, calculated from panels C, F, I, and L. (B-D) 30hr post sync, n = 24; 42hr post sync, n = 19. (E-G) 30hr post sync, n = 33; 42hr post sync, n = 38. (H-J) 30hr post sync, n = 50; 42hr post sync, n = 54. (K-M) CTRL siRNA, 30hr post sync, n = 5; CTRL siRNA, 42hr post sync, n = 4; Herp siRNA, 30hr post sync, n = 11; Herp siRNA, 42hr post sync, n = 5. Values in graphs are mean ± SEM (*p < 0.05, ****p < 0.00005; (D, G, J) t-test, (M) one-way ANOVA. (N-O) Cells were harvested at the indicated times and processed for Western blot analysis. Vinculin and GAPDH served as loading controls for ITPR and BMAL1, respectively. (N) Representative Western blot images from six independent experiments. (O) Densitometric quantification of Western blot data showing relative levels of ITPR1 and ITPR2 at different times. Values in graphs are mean ± SEMs (*p < 0.05, ****p < 0.00005; t-test).

To confirm that HERP underlies this circadian phase-dependent ER Ca2+ release, we compared ER Ca2+ release in control and Herp-KD astrocytes at 30hr and 42hr post sync. Control astrocytes exhibited time-dependent variation in ER Ca2+ release, which was abolished in Herp-KD astrocytes (Figure 4K-4L), indicating the involvement of HERP in this regulation.

We then assessed whether ITPR1 and ITPR2 levels varied with time. BMAL1 phosphorylation served as a proxy for circadian phase (Figure 4N). ITPR2 levels showed time-dependent changes opposite to those of HERP, whereas ITPR1 levels remained constant (Figure 4N and 4O). Since Itpr2 mRNA levels did not exhibit a rhythmic pattern (Figure 3-figure supplement 1), the circadian phase-dependent differences in ITPR2 levels are likely driven by HERP-mediated ITPR2 degradation. Consequently, circadian rhythmic ER Ca2+ release is primarily driven by oscillations in ITPR2.

To verify if ER Ca2+ release is regulated by the circadian clock, we analyzed the ER Ca2+ response in Bmal1-/-mouse astrocyte cultures. In WT astrocyte cultures, there was a clear circadian phase-dependent difference in ER Ca2+ release. However, this difference was abolished in Bmal1-/- astrocyte cultures (Figure 5A-C). The absence of rhythmic ER Ca2+ release in Bmal1-/- astrocytes correlated with constant levels of HERP (Figure 5D-E). Additionally, ER Ca2+ release was significantly faster in Bmal1-/- astrocytes compared to WT astrocytes (Figure 5A-C) attributed to lower levels of HERP in Bmal1-/- astrocytes (Figure 5D-E). Collectively, these results indicate that ER Ca2+ release is regulated by HERP, which is under circadian clock control.

Time-dependent ER Ca2+ release is abolished in cultured astrocytes from Bmal1-/-mice.

Cultured astrocytes Bmal1-/- mice and WT littermates were transfected with G-CEPIA1er and then their circadian rhythm was synchronized by SS. At the indicated Time, astrocytes were treated with 100 μM ATP and Ca2+ imaging was performed. (A) Representative time-lapse images of ER Ca2+ indicator. (B) ΔF/F0 values over time following ATP application. (C) Area above the curve values, calculated from panel B. (A-C) WT 30hr post sync, n = 17; WT 42hr post sync, n = 15; KO 30hr post sync, n = 19; KO 42hr post sync, n = 21. Values are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; one-way ANOVA). (D, E) Cultured astrocytes from Bmal1-/-mice and WT littermates were synchronized by SS. Cells were harvested at the indicated times and processed for Western blot analysis. (D) Representative Western blot images from six independent experiments. GAPDH served as a loading control. (E) Values are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; two-way ANOVA).

ATP-induced S368-phosphorylation of CX43 and gap junctional communication shows circadian variation

Syncytial coupling through gap junctions is a prominent feature of astrocytes, critical for their homeostatic functions, including the diffusion and equilibration of ions, metabolites, and signaling molecules (Langer et al., 2012; Verkhratsky & Nedergaard, 2018). This coupling allows the astrocytic syncytium to propagate Ca2+ waves, affecting nearby and remote cells through release of “gliotransmitters” and modulating synaptic functions in a far-reaching network (Pacholko et al., 2020). Connexin (Cx) family proteins constitute gap junction channels (GJCs) (Giaume et al., 2021), with Cx43 (also known as Gja1) and Cx30 (also known as Gjb6) being the main connexins in astrocytes (Kunzelmann et al., 1999; Nagy et al., 1999). In our cultured astrocytes, Cx43 was the most abundantly expressed connexin (Figure 6-figure supplement 1). Phosphorylation of CX43 at Ser368 by protein kinase C (PKC), activated in response to intracellular Ca2+, is known to decrease gap junction conductance (Enkvist & McCarthy, 1992; Nimlamool et al., 2015; Solan & Lampe, 2014). Accordingly, we hypothesized that the circadian variations in ATP-driven ER Ca2+ responses might differentially impact CX43 phosphorylation according to times of day.

First, we confirmed that ATP treatment induces rapid phosphorylation of CX43 at Ser368 (pCX43 (S368)) in cultured astrocytes (Figure 6A and 6B). We then examined whether pCX43 (S368) varies with circadian phase. Notably, ATP-induced ER Ca2+ release was more pronounced at 30hr than at 42hr post sync (Figure 3), and pCX43 (S368) in response to ATP was also significantly higher at 30hr compared to 42hr post sync (Figure 6C and 6D).

ATP-induced pCX43 (S368) levels varies according to time post sync.

(A-D, G, H) Cultured astrocytes with (C, D) or without (A, B) synchronization, were treated with 100 µM ATP either at 30hr post sync or 42hr post sync and processed for Western blot analysis at the indicated times. Vinculin, GAPDH, and/or β-tubulin (TUBB) served as loading controls. The intensity of pCX43 (S368) and Cx43 for each sample was normalized to that of Vinculin. (A) Representative Western blot images from more three independent experiments. Vinculin served as a loading control. (B) Densitometric quantification of Western blot data, showing relative pCX43 (S368)/CX43 levels. Values are normalized to those for mock-treated samples at time zero (set to 1). (C) Representative Western blot images from three independent experiments. Vinculin, GAPDH, and β-tubulin (TUBB) served as loading controls. (D) Densitometric quantification of Western blot data, showing relative pCX43 (S368)/CX43 levels. Values are normalized to those for mock-treated samples at ATP zero for 30hr post sync (set to 1). (E, F) Changes in CX43 phosphorylation in vivo. Mice were entrained to a 12-hour light/dark cycle followed by constant dark conditions. At the indicated times, the prefrontal cortex area was dissected and processed for Western blot analysis. (E) Representative Western blot images from three independent experiments. (F) Densitometric quantification of Western blot data showing relative levels of pCX43 (S368). (G, H) 48hrs post siRNA transfection, cells were synchronized by serum shock. At the indicated times post sync, 100 μM ATP was treated and processed for Western blot analysis (G) Representative Western blot images from three independent experiments. Vinculin served as a loading control. (H) Densitometric quantification of Western blot data, showing relative pCX43 (S368)/CX43 levels. Values are normalized to those for 30hr post sync at ATP zero (min) (set to 1). Values are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, ****p < 0.00005; (B, D, H) two-way ANOVA, (F) t-test).

We also examined circadian variations of pCX43 (S368) in vivo. Mice were entrained to a 12:12 hour light/dark (LD) cycle for two weeks and then kept in constant darkness (DD). This ensured that variations in CX43 phosphorylation were due to the internal circadian clock and not external light stimuli. Prefrontal cortex tissues were dissected at circadian time (CT) 58 and CT70 on the third day of DD. pCX43 (S368) levels were higher at CT70 (subjective night) than at CT58 (subjective day), consistent with our in vitro results (Figure 6E and 6F).

To confirm that HERP underlies the rhythmic phosphorylation of CX43 at Ser368, we compared pCX43 (S368) in control and Herp-KD astrocytes at 30hr and 42hr post sync. In control astrocytes pCX43 (S368) was higher at 30hr compared to 42hr, consistent with previous results. However, in Herp-KD astrocytes, the circadian phase-dependent differences in pCX43 (S368) were abolished despite the normal circadian rhythm indicated by BMAL1 phosphorylation, (Figure 6G and 6H). This suggests that rhythms of HERP levels are required for the rhythmic phosphorylation of CX43 (S368).

Finally, we assessed if rhythmic phosphorylation of CX43 contributes to circadian phase-dependent variation in gap junctional communication in cultured astrocytes by using fluorescence recovery after photobleaching (FRAP) method. The absence of FRAP signal in carbenoxolone (CBX)-treated astrocytes confirmed the involvement of gap junctions in intercellular communication of cultured astrocytes (Figure 7A-7C). We observed a faster and more pronounced recovery of fluorescence in photobleached astrocytes at 42hr compared to 30hr post sync (Figure 7D-7F). Importantly, this variation was absent in astrocytes cultured from Bmal1-/- mice, indicating that circadian clock regulates gap junctional communication (Figure 7G.-7I).

Gap junction conductance varies according to time post sync.

(A-C) 20μM carbenoxolone (CBX), a gap junction channel block, was applied to cultured astrocytes for 30 minutes before gap-FRAP analysis. (A) Representative time-lapse images of prebleaching, bleaching and recovery condition during gap-FRAP analysis. (B) Ft/F0 values over time following photobleaching (yellow rectangle). (C) Recovery % values, calculated from panel B. Mock, n = 7; CBX, n = 7. Values in graphs are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; Mann-Whitney U test). (D-I) Cultured astrocytes from WT (D-F) and Bmal1-/- (G-I) mice were synchronized by SS. At the indicated time, gap-FRAP was performed. (D, G) Representative time-lapse images of prebleaching, bleaching and recovery condition during gap-FRAP analysis at 30hr post sync and 42hr post sync. (E, H) Ft/F0 values over time following photobleaching (yellow rectangle). (F and I) Recovery % values, calculated from panel H and L, respectively. (D-F) 30hr post sync, n = 15; 42hr post sync, n = 11. (G-I) 30hr post sync, n = 9; 42hr post sync, n = 9. Values in graphs are mean ± SEM (*p < 0.05, **p < 0.005, ***p < 0.0005, and ****p < 0.00005; F : t-test, I : Mann-Whiteny U test). (J) Schematic diagram illustrating the regulation of ER Ca2+ response by the circadian clock through rhythmic oscillation of HERP. Refer to the text for the detailed explanation.

In summary, we propose a model illustrating how the circadian clock influences astrocyte function through the regulation of ER Ca2+ response. Based on phase comparisons of clock genes between our data and CircaDB (Figure 1D), we defined subjective day and night, with 8hr post sync corresponding to the beginning of the day. In astrocytes, BMAL1/CLOCK-controlled oscillation of HERP regulates day and night variations in ITPR2 levels. These oscillations contribute to changes in ER Ca2+ responses, which in turn result in the distinct day and night differences in pCx43 (S368) and the gap junction conductance (Figure 7J).

Discussion

It is widely acknowledged that the circadian clock orchestrates cellular functions in a highly cell type-specific manner by governing transcription (Mure et al., 2018; Zhang et al., 2014). Although the importance of the circadian clock in astrocytes has been well-documented (Barca-Mayo et al., 2020; Barca-Mayo et al., 2017; Griffin et al., 2019; Lananna et al., 2018; Prolo et al., 2005; Tso et al., 2017), studies specifically exploring astrocyte-specific circadian transcriptomes have been limited. Our study conducted such an analysis on primary cultured cortical astrocytes, discovering that rhythmic expression of Herp underlies day/night variation in ER Ca2+ response by regulating IP3R levels. We also found that Ca2+-activated CX43 phosphorylation varied according to the daily rhythmic ER Ca2+ response. Given that astrocytes function in a network in which Ca2+ propagates through GJCs, we propose that the time-dependent control of ER Ca2+ responses could influence the role of astrocytes in modulating synaptic activity throughout the day (see below).

Circadian oscillations of Herp expression have been previously documented in various mouse tissues including the lung, heart, liver, kidney, aorta, skeletal muscle, adrenal gland, pancreatic islet and SCN (Panda et al., 2002; Rakshit et al., 2016; Zhang et al., 2014). In this study, we identified that Herp is also rhythmically expressed in cultured astrocytes. Herp exhibited rhythmic expression in phase with Per2 and showed flattened expression at its nadir level in Bmal1-/- astrocyte cultures, suggesting direct regulation CLOCK/BMAL1 heterodimers (Figure 2A). ChIP-Atlas analysis (https://chip-atlas.org) (Oki et al., 2018; Zou et al., 2022) has shown CLOCK binding to the genomic region of Herp (Annayev et al., 2014; Hong et al., 2018). Although no canonical E-Box is present in the upstream 6kb region of the Herp gene, several non-canonical E-Boxes, important for circadian expression (Yoo et al., 2005; Zhang et al., 2011), were identified (Table S2). Among these, the non-canonical E-Box sequence “CACGTT,” located near transcription start site (chr8:94386194-94386543), is highly conserved across diverse mammalian species. Therefore, we propose that these non-canonical E boxes might drive the CLOCK/BMAL1 dependent expression of Herp. A recent study employing a circadian single-cell RNA seq analysis of SCN reported that Herp displays a circadian rhythm exclusively in astrocytes and oligodendrocytes (Wen et al., 2020). Given that astrocytic Ca2+ levels in the SCN show a circadian rhythm, peaking during the subjective night and are crucial for circadian timing (Brancaccio et al., 2019; Patton et al., 2022), HERP’s circadian rhythmic expression in SCN astrocytes (Wen et al., 2020) may be integral to the mechanisms of circadian timing.

We demonstrated that HERP regulated ER Ca2+ responses by degrading IP3Rs in astrocytes (Figure 3J-3M). (Paredes et al., 2016; Torrealba et al., 2017). When Herp was downregulated, both ITPR1 and ITPR2 levels increased, suggesting HERP’s role in degrading both ITPR1 and ITPR2. However, only the levels of ITPR2 exhibited circadian phase-dependent variation (Figure 3J-3M, Figure 4K-4L). The min/max ratio of HERP expression throughout the day was 0.38 under normal conditions but 0.072 in Herp-KD conditions. Based on this, we reasoned that the amplitude of HERP oscillations impacted the levels of ITPR2 under normal circumstances, while having a much lesser effect on the levels of ITPR1. Thus, the circadian control of ER Ca2+ release in response to stimulation in cultured astrocytes is primarily attributable to ITPR2.

ER and cytosolic Ca2+ responses following ATP treatment exhibited distinct circadian variation, but mitochondrial Ca2+ responses did not show significant variations (Figure 4). We think of several possibilities to account for the absence of significant circadian variations in mitochondrial Ca2+ relative to cytoplasmic Ca2+ change. Upon ATP stimulation, direct Ca2+ transfer occurs through IP3R in the cytosol. For mitochondrial uptake, Ca2+ first transits to the intermembrane space via IP3Rs and the voltage-dependent anion channel (VDAC) complex before entering the matrix through mitochondrial calcium uniporter (MCU) (Giorgi et al., 2018). The Mito-R-GECO1 reporter is confined within the mitochondrial matrix and thus limiting assessment of Ca2+ levels in the intermembrane space (Wu et al., 2013). Substantial changes in Ca2+ levels might occur in the intermembrane space, but the influx into the matrix via MCUs could remain constant across different time, resulting in apparent lack of significant change. Secondly, the amount of VDAC or the VDAC:IP3R complex on the mitochondrial membrane might vary inversely with HERP levels, potentially maintaining a consistent flow of Ca2+ into the mitochondria irrespective of the time. Additionally, the presence of several Ca2+ extrusion mechanisms in mitochondria (Giorgi et al., 2018) points to a more complex regulation of Ca2+ compared with that in the cytosol. Although the precise mechanism remains to be elucidated, our data indicate that the circadian variation in Ca2+ release from the ER, particularly in response to stimuli, predominantly impacts cytoplasmic signaling rather than mitochondrial signaling.

During the subjective night phase (active period for mice), ATP- stimulated ER Ca2+ release was pronounced than during the subjective day phase (rest period for mice) (Figure 4). This finding aligns with studies showing heightened stimulus-evoked Ca2+ activity in astrocytes during wakefulness, which diminishes during sleep (Bojarskaite et al., 2020; Ingiosi et al., 2020). In flies, Ca2+ activity in astrocytes is correlated with the duration of wakefulness, suggesting that astrocytic Ca2+ is a key factor in the homeostatic regulation of sleep (Blum et al., 2021). Our findings provide a mechanistic explanation for differential Ca2+ levels observed between wake and sleep states, with circadian regulation of ITPR2 by HERP being a key factor in this process.

ATP-stimulated Cx43 (S368) phosphorylation is higher at 30hr (subjective night phase) than at 42hr post sync (subjective day phase) (Figure 6C and 6D.), a finding further supported by in vivo experiments (Figure 6E and 6F). What are the implications of this day/night variation in CX43 (S368) phosphorylation? We reasoned that the circadian variation in CX43 phosphorylation could significantly impact astrocyte functionality within the syncytium. Indeed, our cultured astrocytes exhibited circadian phase-dependent variation in gap junctional communication (Figure7D-7F). Astrocytes influence synaptic activity through the release of gliotransmitters such as glutamate, GABA, D-serine, and ATP, triggered by increases in intracellular Ca2+ in response to the activity of adjacent neurons and astrocytes (Verkhratsky & Nedergaard, 2018). Importantly, this increase in Ca2+ spreads to adjacent astrocytes through GJCs (Fujii et al., 2017), influencing a large area of the neuronal network. Considering that Cx43 Ser368 phosphorylation occurs to uncouple specific pathways in the astrocytic syncytium to focus local responses (Enkvist & McCarthy, 1992), our findings suggest that astrocytes better equipped for localized responses when presented with a stimulus during the active phase in mice. Conversely, during the rest period, characterized by more synchronous neuronal activity across broad brain areas (Vyazovskiy et al., 2009) higher GJC conductance might allow astrocytes to exert control over a larger area. In support of this idea, recent study showed that synchronized astrocytic Ca2+ activity advances the slow wave activity (SWA) of the brain, a key feature of non-REM sleep (Szabó et al., 2017). Blocking GJC was found to reduce SWA, further supporting this interpretation. However, conflicting findings have also been reported. For instance, Ingiosi et al. (Ingiosi et al., 2020) found that astrocytic synchrony was higher during wakefulness than sleep in the mouse frontal cortex. Whether these differing results in astrocyte synchrony during resting and active periods are attributable to differences in experimental context (e.g., brain regions, sleep-inducing condition) remains unclear. Indeed, astrocyte Ca2+ dynamics during wakefulness/sleep vary according to brain regions (Tsunematsu et al., 2021). While the extent of astrocyte synchrony might differ depending on brain region and/or stimulus, our results suggest that the baseline state of astrocyte synchrony, which is affected by GJC conductance, varies with the day/night cycle.

In conclusion, our findings enhance the understanding of the circadian regulation of astrocytic functions, particularly through the rhythmic expression of HERP and its impact on ER Ca²⁺ signaling and CX43 phosphorylation. This study underscores the intricate interplay between the circadian clock and astrocyte physiology, highlighting the pivotal role of HERP in modulating key astrocytic processes such as sleep, gap junction communication and synaptic modulation. The day/night variations in ER Ca²⁺ response and CX43 phosphorylation, and their implications for astrocytic network dynamics and neuronal interactions, open new avenues for exploring the broader implications of astrocytic circadian rhythms in brain function and health.

Materials and Methods

Animals

All procedures for the care and use of laboratory animals were approved by the Institutional Animal Care and Use Committee (IACUC) of Ajou University School of medicine. Bmal1-/+ mice (B6.129-Bmal1tm1Bra/J; Jackson Laboratory, #009100)(Bunger et al., 2000) were purchased from Jackson Laboratory(USA) and housed in a specific pathogen-free environment at the Animal Research Center of Ajou University Medical Center, with a standard 12-hour light/dark cycle and free access to food. Primary cultured astrocytes involving only WT mice used 1day-old C57BL/6 mice (Koatch Inc., Korea). In experiments employing primary cultured astrocytes from Bmal1-/- mice, wild type (WT) littermates were used as controls. For animal studies, 7-8 week-old adult C57BL/6 male mice were first entrained in a standard 12-hour light/dark cycle for 2 weeks and then maintained in a constant darkness (DD). On the third day of DD, mice were sacrificed at indicated time and its prefrontal cortex (PFC) was dissected.

Primary astrocyte culture

Primary astrocyte culture was prepared according to a previous report (Choi et al., 2018) with minor modifications. In brief, 1 day old pups were anesthetized by hypothermia and their cerebral cortices were dissected out and triturated in Modified Eagle’s medium (MEM; Welgene, Korea, LM 007-11) containing 10% fetal bovine serum (FBS; Hyclone, USA, SV30207.02), penicillin/streptomycin (Gibco, USA, 15140-122), 10 µM HEPES (Gibco, USA, 15630080) and GlutaMAX (Gibco, USA, 35050-061) yielding a single-cell suspension. Cells were plated into 75 cm2 T-flasks (1pup/flask) and incubated at 37°C in a humidified 5% CO2 incubator for 2 weeks. Primary astrocytes were then incubated in a serum-free MEM for 7 days, after which astrocytes were detached from the T-flask with 0.25% trypsin and plated on culture dishes (1.0X106cells/60 mm dish) for experiments.

Circadian transcriptome analysis

The experimental scheme for synchronizing circadian clocks of cultured astrocyte is illustrated in Figure 1A. In brief, astrocytes were grown in dishes (1.0 X 106cells/60 mm dish) until reaching complete confluence and then incubated in serum-free MEM for 72hrs. Cells were then subjected to serum shock (SS), a well-established procedure for synchronizing the circadian clock of cultured cells (Balsalobre et al., 1998), by exchanging the medium for MEM containing 50% horse serum. 12hrs post SS, astrocytes were harvested at 4-h intervals for 2 days for RNA sequencing (RNA-seq). Total RNA was extracted and purified from harvested astrocytes using a RNeasy Plus Micro Kit (Qiagen, Germany, 74034). RNA quality and quantity assessments, RNA-Seq library construction, and next-generation sequencing analysis were performed by Macrogen Inc. (Korea). Total RNA integrity and library size were analyzed using an Agilent Technologies 2100 Bioanalyzer. Paired-end raw reads, generated using an Illumina HiSeq 2000 system, were aligned to the mm10 mouse genome using STAR aligner (v.2.5.2b)(Dobin et al., 2013). Gene expression was subsequently quantified by calculating transcripts per million (TPM) using Ensemble gene (release 82) annotations. The threshold for defining expressed transcripts was set using a Gaussian mixture model in the R package ‘mixtools’. To detect circadian oscillating transcripts, we employed two circadian oscillation detection methods BioCycle (Agostinelli et al., 2016) and MetaCycle (Wu et al., 2016).

Quantitative reverse transcription polymerase chain reaction (qRT-PCR)

Total RNA was extracted from cells and purified using the RNeasy Plus Micro Kit (Qiagen, 74034). 1μg of total RNA was reverse transcribed using an oligo-dT primer and PrimeScript RTase (TaKaRa, Japan, 2680A). Quantitative real-time PCR was performed using a Rotor Gene Q (Qiagen) with TB Green Premix Ex Taq (Takara, RR420A) The specific primers used were provided in the Table S3. Noncycling mRNA encoding HPRT was used to normalize gene expression. The data were analyzed using Rotor Gene 6000 software, and the relative mRNA levels were quantified using the 2-ΔΔCt method in which ΔΔCt=[(Ct target - Ct HPRT) of experimental group]-[(Ct target - Ct HPRT) of control group].

siRNA transfection and Immunoblotting

The following On-Target plus SMARTpool siRNAs were purchased from Dharmacon (USA): Non-targeting CTRL siRNA (D-001810-01-50), Herp siRNA (L-049714-01-0005), Itpr1 siRNA (L-040933-00-0005), and Itpr2 siRNA (L-041018-00-0005). Upon reaching >90% confluence, the medium was replaced with serum-free MEM, and cells were incubated for 72hrs. Cells were then transfected with siRNA using Lipofectamine RNAiMAX (ThermoFisher, USA, 13778150) per the manufacturer’s instructions. Experiments were conducted 48hrs post-transfection.

For immunoblotting, astrocytes were lysed using modified-RIPA buffer (50 mM Tris-HCl pH 7.4, 1% NP-40, 0.5% sodium deoxycholate, 150 mM NaCl). Mouse prefrontal cortex was lysed using T-Per (ThermoFisher, 78510) with protease inhibitor cocktail (Sigma-Aldrich, P8340) and phosphatase inhibitor cocktails 2 and 3 (Sigma-Aldrich, P5726 and P0044). Proteins were separated by SDS-PAGE and transferred to polyvinylidene fluoride membranes. Membranes were blocked with 5% skim milk and incubated overnight at 4°C with primary antibodies: anti-BMAL1 (Abcam (UK), ab93806), 1:2000; anti-HERP (Abcam, ab150424), 1:1000; anti-ITPR1 (Abcam, ab264284), 1:1000; anti-ITPR2 (Alomone Labs, Israel, ACC-116), 1:1000; anti-CX43 (Sigma, C6219), 1:5000; anti-pCX43 (Ser368) (CST, USA, 3511), 1:1000; anti-GAPDH (Novus, USA, NB100-56875), 1:5000; anti-vinculin (Sigma-Aldrich, V4505), 1:5000; and anti-total ERK (CST, 9102), 1:5000. Membranes were washed with TBST, incubated with secondary antibodies, and visualized using enhanced chemiluminescence. Protein levels were quantified by densitometric analysis of band intensities using ImageJ software.

Ca2+ reporter plasmid transfection and imaging analysis

After cells reached >90% confluence, the medium was replaced with serum-free MEM. For Ca2+ measurements, cells were transfected with Ca2+ indicator plasmids pCMV-G-CEPIA1er (a gift from Masamitsu Iino, Addgene plasmid #58215) (Suzuki et al., 2014), CMV-mito-R-GECO1, or CMV-R-GECO1 (gifts from Robert Campbell, Addgene plasmid #46021, #32444) (Wu et al., 2013) 72hrs post-serum deprivation using Lipofectamine 2000, with or without siRNA treatment. For Xestospongin C (XesC) experiments, 10 μM XesC was applied 30 minutes before imaging. 48hrs post-transfection, the medium was replaced with Ca2+- and Mg2+-free HBSS (Gibco, 14175095). Fluorescence images were captured every 3 seconds using a Nikon A1R Confocal Microscope with a x60 1.4 NA Plan-Apochromat objective at 37°C and 5% CO2. These experiments were conducted at the Three-Dimensional Immune System Imaging Core Facility of Ajou University (Korea). Images were analyzed using NIS Elements C software. Fluorescence intensity at each time point (Ft) and ΔF values (Ft-F0) were calculated. F0 values were averaged from the fluorescence intensity of 10 frames prior to stimulation.

Ca2+ reporter plasmid co-localization analysis

After cells reached >90% confluence, the medium was replaced with serum-free MEM. 72hrs later, cells were transfected with either the ER Ca2+ reporter plasmid pCMV-G-CEPIA1er or the mitochondrial Ca2+ reporter CMV-mito-R-GECO1. For ER co-localization analysis, pCMV-G-CEPIA1er was co-transfected with DsRed2-ER-5 (a gift from Michael Davidson, Addgene plasmid #55836) (Day & Davidson, 2009). For mitochondrial co-localization, 200 nM Mitotracker was applied 30 minutes before imaging. Live imaging was performed using a Nikon A1R Confocal Microscope with a x60 1.4 NA Plan-Apochromat objective at 37°C and 5% CO2. These experiments were conducted at the Three-Dimensional Immune System Imaging Core Facility of Ajou University (Korea). Images were analyzed using NIS Elements C software.

Gap-FRAP (fluorescence recovery after photobleaching) analysis

Astrocyte communication was evaluated using a modified gap-FRAP assay (Santiquet et al., 2012). 30 minutes before imaging, cells were incubated with 0.5 μM Calcein-AM (Invitrogen, C1430) and 0.05% Pluronic-127 (Invitrogen, P6866) at 37°C. After incubation, cells were rinsed twice with serum-free MEM to remove excess dye. For carbenoxolone (CBX) experiments, 20 μM CBX was added for 30 minutes after calcein incubation. Immediately before imaging, the medium was replaced with Ca2+- and Mg2+-free HBSS (Gibco, 14175095). Live imaging was performed using a Nikon A1R Confocal Microscope with a 488 nm argon laser and a x60 1.4 NA Plan-Apochromat objective. Photobleaching was performed on a region of interest (ROI) cell, and fluorescence was measured at 37°C and 5% CO2. Fluorescence intensity was recorded pre-bleach (5 images), during a 15-second laser pulse (100% power, 5.3 lines/sec), and post-bleach (every 3 seconds for 3 minutes). Images were analyzed using NIS Elements C software. Fluorescence intensity at each time point (Ft) was normalized to the pre-bleach intensity (F0), calculated as the average of 5 pre-bleach frames. The percentage of fluorescence recovery was calculated using: (FFR-FA)/(F0-FA)x100, where FFR is the full recovery fluorescence intensity and FA is the intensity immediately after photobleaching.

Quantification and statistical analysis

To conduct the statistical analyses, all samples were first tested for normality. For cases where normality testing was not possible, such as with the Area Under/Above Curve, t-tests were used for comparing two groups, and one-way ANOVA was used for comparing three or more groups, considering the sufficient sample size. When comparing two groups, a t-test was performed if the data followed a normal distribution; otherwise, the Mann-Whitney U test was used. For comparisons among three or more groups, one-way ANOVA was used if the data were normally distributed, and the Brown-Forsythe and Welch ANOVA tests were used if they were not. When comparing three or more groups across multiple time points, a two-way ANOVA was performed. A p < 0.05 was considered statistically significant for all tests.

Data and code availability

Circadian RNA sequencing data have been deposited at GEO under accession number GEO: GSE254678.

Any additional information required to reanalyze the data reported in this paper is available from the E.Y.K upon request.